\documentclass[../notes.tex]{subfiles}

\pagestyle{main}
\renewcommand{\chaptermark}[1]{\markboth{\chaptername\ \thechapter\ (#1)}{}}
\setcounter{chapter}{1}

\begin{document}




\chapter{Intro to Reactions and Ligands}
\section{Lecture 4: Substitution Reactions}
\begin{itemize}
    \item \marginnote{4/5:}Association/dissociation reactions.
    \item Fairly related to organic S\textsubscript{N}2 and S\textsubscript{N}1 reactions, respectively.
    \item General form:
    \begin{equation*}
        \ce{ML6 + L$'$ <=> ML5L$'$ + L}
    \end{equation*}
    \item We investigate the position of the equilibrium with the three main characteristics that determine reactivity.
    \begin{enumerate}
        \item Sterics.
        \begin{itemize}
            \item Related to the metal coordination number.
            \begin{itemize}
                \item $\text{C.N.}>6$ is typically disfavored.
                \item $\text{C.N.}<6$ is possible.
            \end{itemize}
            \item The size of \ce{L$'$} is also important: If $\ce{L$'$}=\ce{PPh3}$ for example, this is hard to get to $\text{C.N.}>4$.
        \end{itemize}
        \item Ligand character.
        \begin{itemize}
            \item In nonpolar media, dissociation of charged groups (e.g., \ce{Cl-}) will be disfavored. However, the opposite is true in polar media.
            \begin{itemize}
                \item This is because of the issue of making charge/ionizing.
            \end{itemize}
            \item The match between \ce{M} and \ce{L} (e.g., hard/soft, electron rich/poor) is also important.
            \begin{itemize}
                \item For example, \ce{Fe^0} will bind \ce{CO} strongly since \ce{Fe^0} is electron rich and \ce{CO} is a $\pi$ acceptor.
                \item However, \ce{Fe^{IV}} will not (as a hard, electron-poor metal center).
            \end{itemize}
        \end{itemize}
        \item Electronic structure of the metal center (whether or not the metal is electronically saturated [has 18 electrons]).
        \begin{itemize}
            \item $18\,\e[-]$: it will not want to coordinate an additional \ce{L$'$}.
            \item $20\,\e[-]$: it will want to dissociate.
            \item $16\,\e[-]$: it \emph{can} associate.
            \begin{itemize}
                \item However, it may not want to given that $16\,\e[-]$ square-planar complexes are fairly stable.
                \item The associated state may be a transition state in a square-planar ligand substitution or otherwise not a ground state.
            \end{itemize}
        \end{itemize}
    \end{enumerate}
    \item Ligand substitution reactions terms: \textbf{Kinetic} and \textbf{thermodynamic}.
    \item \textbf{Kinetic} (considerations): Elements are inert (slow) or labile (fast).
    \item \textbf{Thermodynamic} (considerations): Which side of an equilibrium will be favored. Elements are stable or reactive.
    \item In ligand substitution reactions, there are two limiting regimes:
    \begin{enumerate}
        \item Associative substitution.
        \begin{itemize}
            \item See the related discussion in \textcite{bib:CHEM20100Notes}.
            \item This is the most general reaction type, even for coordinatively saturated complexes.
            \item Rate law:
            \begin{equation*}
                \dv{\ce{[ML5L$'$]}}{t} = k_\text{obs}\ce{[ML6][L$'$]}
            \end{equation*}
        \end{itemize}
        \item Dissociative mechanism.
        \begin{itemize}
            \item See the related discussion in \textcite{bib:CHEM20100Notes}.
            \item There are many things that look dissociative that are associative (e.g., instead of forming a 5-coordinate species, you could just have a molecule of the solvent displace a ligand).
            \item This mechanism is rare and hard to prove.
            \item Rate law:
            \begin{equation*}
                \dv{\ce{[ML5L$'$]}}{t} = \frac{k_2k_1\ce{[ML6][L]}}{k_{-1}\ce{[L]}+k_2\ce{[L$'$]}}
            \end{equation*}
            \item Experimentally, we swamp the reaction with \ce{L$'$} so that $\ce{[L$'$]}>>>$ than all other reagents. This makes it so that the rate is just $k_\text{obs}\ce{[ML6]}$, i.e., pseudo-first order conditions.
        \end{itemize}
    \end{enumerate}
    \item Unfortunately, much like in orgo, very few cases are at these extremes and we can have hybrids called\dots
    \begin{enumerate}[resume]
        \item Interchange mechanisms.
        \begin{itemize}
            \item See the related discussion in \textcite{bib:CHEM20100Notes}.
            \item Within this category, we can have $I_a$ (associative interchange) and $I_d$ (dissociative interchange).
            \item In the transition state, we have \ce{L$'$} coming in and \ce{L} leaving at the same time.
        \end{itemize}
    \end{enumerate}
    \item Kinetics and rates of these mechanisms.
    \item Several categories (measure with water exchange rates; see \textcite{bib:CHEM20100Notes}):
    \begin{enumerate}[label={\Roman*)}]
        \item Very fast.
        \begin{itemize}
            \item Alkali metals (species that primarily engage in ionic bonding; little covalent character).
            \item $\SI{e8}{\per\second}$; close to the diffusion limit.
        \end{itemize}
        \item Fast.
        \begin{itemize}
            \item Higher valent ions; often \ce{M^3+} such as \ce{Al^3+}.
            \begin{itemize}
                \item Higher charge $\Rightarrow$ higher ligand affinity $\Rightarrow$ slightly slower but still pretty fast.
            \end{itemize}
            \item $\num{e3}$-$\SI{e8}{\per\second}$.
        \end{itemize}
        \item Slower.
        \begin{itemize}
            \item Getting into the transition metals: \ce{Fe^3+}, \ce{V^3+}, \ce{Ti^3+}.
            \begin{itemize}
                \item $d$-orbital splitting $+$ covalency $\Rightarrow$ stronger bonding $\Rightarrow$ slower exchange rate.
            \end{itemize}
            \item $\num{e1}$-$\SI{e4}{\per\second}$.
        \end{itemize}
        \item Inert.
        \begin{itemize}
            \item \ce{Co^3+}, \ce{Cr^3+}, \ce{Pt^2+}, and \ce{Fe^2+}(L.S.).
            \item $\num{e-8}$-$\SI{e-4}{\per\second}$.
        \end{itemize}
    \end{enumerate}
    \item The overlap between the rates reflects the fact that there is no hard and fast cut off between categories.
    \item The identity of \ce{L$'$} also influences rates.
    \begin{itemize}
        \item Reaction rates increase with the ligand field strength of \ce{L$'$}\footnote{Goes over Table IX.1 from \textcite{bib:CHEM20100Notes}.}.
    \end{itemize}
    \item Characteristics of the metal that control the observed rates.
    \begin{itemize}
        \item Ranking L.S. metal centers (slowest to fastest): $\ce{Co^{III}}<\ce{Cr^{III}}<\ce{Mn^{III}}<\ce{Fe^{III}}<\ce{Ti^{III}}<\ce{V^{III}}$.
        \item Considering the $d$ counts, we have $d^6<d^3<d^4<d^5<d^1<d^2$.
        \item Now think of this in terms of the $d$-orbitals splitting diagram (Figure \ref{fig:6-octab}).
        \begin{itemize}
            \item As the antibonding orbitals get filled, $\sigma$ bonds will weaken, promoting a faster exchange.
            \item Full and half-full $t_{2g}$ also provides stability.
        \end{itemize}
    \end{itemize}
    \item Thus, we list the following configurations as inert and labile (see the related discussion in \textcite{bib:CHEM20100Notes}):
    \begin{itemize}
        \item Inert: $d^3$, L.S. $d^{4,5,6}$, and square planar $d^8$.
        \item Labile: $d^0$, $d^1$, $d^2$, H.S. $d^{4,5,6}$, $d^7$, $d^9$, $d^{10}$.
    \end{itemize}
    \item Other important kinetic factors:
    \begin{enumerate}
        \item Oxidation state.
        \begin{itemize}
            \item As oxidation state increases, exchange rate decreases (becomes more inert).
        \end{itemize}
        \item Size.
        \begin{itemize}
            \item Smaller ions are more inert.
            \item However, first row ions are almost always labile (because they more readily populate higher spin states).
        \end{itemize}
        \item Chelate effect.
        \begin{itemize}
            \item Reviews some info from \textcite{bib:CHEM20100Notes}.
            \item Chelating ligands form a ring or a \textbf{metallacycle} (this is why 4,5-membered ligands are stable; because 5,6-membered rings are favorable).
            \item Binding of a chelating ligand is typically favored, primarily due to entropic reasons (effective concentration is secondary).
            \item Example: Gives actual $\Delta G=\Delta H-T\Delta S$ thermodynamic data for the formation reaction of \ce{Cu(MeNH2)4^2+} vs. \ce{Cu(en)2^2+} to emphasize the importance of entropy (see the related discussion in the notes on Chapter 10 in \textcite{bib:CHEM20100Notes}).
            \item EDTA is a hexadentate ligand that is commonly used in biology to pull all metal centers out of solution.
            \begin{itemize}
                \item For \ce{Fe^3+} for example, $K_f=\SI{e25}{\per\mole}$. What is $\si{\per\mole}$ and why is it here?
                \item Sidenophones and euterobactin are biology's own chelaters ($K_f=\SI{e52}{\per\mole}$).
                \item These chelaters involved because if bacteria are going to invade a host, they need to scavenge iron, but iron is pretty tightly regulated. Thus, there has been an arms race of molecules that can scavenge iron or prevent iron from being scavenged.
            \end{itemize}
            \item Chelation therapy: If exposed to a heavy metal, you will be given chelating agents that will bind to metal ions and cause them to be excreted from the body.
        \end{itemize}
        \item Trans effect.
        \begin{itemize}
            \item Reviews some info from \textcite{bib:CHEM20100Notes}.
            \item Helps predict the \textbf{regiochemistry} of where a given ligand will substitute.
            \item Cis-platin reaction mechanism: \ce{{\emph{cis}-}Pt(NH3)2(Cl)2 -> {\emph{cis}-}Pt(NH3)2(H2O)2} in the body, which binds to DNA on the \emph{cis}-water side, causing a kink, stopping transcription, and initiating apoptosis.
            \begin{itemize}
                \item Cis-platin is quite toxic (people are trying to develop formulations that are less so), but highly effective at stopping cancer.
                \item Can't have \emph{trans} because it doesn't have the \emph{cis}-water side. Thus, this synthesis mechanism doesn't work: \ce{[PtCl4]^2- ->[2NH3] {\emph{trans}-}Pt(NH3)2Cl2}.
                \item Therefore, we synthesize it as follows.
                \begin{align*}
                    \ce{K2PtCl4} &\ce{->[4KI]} \ce{PtI4^2-}\\
                    &\ce{->[2NH3]} \ce{{\emph{cis}-}Pt(NH3)2(I)2}\\
                    &\ce{->[1) AgNO3][2) XS KCl]} \ce{{\emph{cis}-}Pt(NH3)2(Cl)2}
                \end{align*}
                \item Note that we start from tetrachloroplatinate because it is the most common form of platinum.
                \item Also note that XS stands for "excess."
            \end{itemize}
            \item Trans-effect order listed.
            \item The trans-effect is kinetic; concerned with rates of exchange.
            \begin{itemize}
                \item Stronger \emph{trans}-directors \textbf{labelize} the ligands opposite them.
            \end{itemize}
            \item The trans influence is thermodynamic.
            \begin{itemize}
                \item It influences the ground state structure, causing lengthening of bonds \emph{trans} to a strong-field ligand (think of this in terms of competition for electrons on the central atom; a strong-field ligand will attract more of these, making the other bond weaker).
            \end{itemize}
        \end{itemize}
    \end{enumerate}
    \item Note that intramolecular reactions (such as a second binding of a bidentate chelating ligand) are highly favored.
\end{itemize}



\section{Lecture 5: Electron Transfer Reactions}
\begin{itemize}
    \item \marginnote{4/7:}More unique to inorganic chemistry since metal atoms have access to many more electrons than common organic atoms.
    \item General form:
    \begin{equation*}
        \ce{M^n+ <=>[$-\e[-]$][$+\e[-]$] M^{(n+1)+}}
    \end{equation*}
    \begin{itemize}
        \item The forward reaction is known as \textbf{oxidation} (metal oxidation state increases), while the reverse is known as \textbf{reduction} (metal oxidation state decreases).
        \item This is different than the oxidation/reduction reactions of organic chemistry, which involve removing or adding, respectively, a hydrogen.
    \end{itemize}
    \item This redox chemistry is important because many transition metals have access to multiple oxidation states.
    \item Two Nobel prizes in this area:
    \begin{itemize}
        \item Henry Taube (1983): Electron transfer in metals.
        \item Rudy Marcus (1992): Marcus theory of electron transfer.
    \end{itemize}
    \item 2 general flavors of electron transfer reactions: \textbf{inner sphere} and \textbf{outer sphere}.
    \item \textbf{Inner sphere}: Bonds are formed.
    \item \textbf{Outer sphere}: No bonds are formed.
    \item Example:
    \begin{itemize}
        \item Consider the reaction \ce{Fe(CN)6^4- + Mo(CN)8^3- -> Fe(CN)6^3- + Mo(CN)8^4-} (electron transfer from iron to molybdenum).
        \item The energies at play: \ce{A^{(n+1)} + B^n -> [A^{(n+1)} + B^n] -> [A^n + B^{(n+1)}]^* -> A^n + B^{(n+1)}}.
        \item Reactants $\to$ encounter complex $\to$ electron transfer state (an excited state) $\to$ products.
    \end{itemize}
    \item Energies:
    \begin{figure}[H]
        \centering
        \begin{subfigure}[b]{0.4\linewidth}
            \centering
            \begin{tikzpicture}
                \small
                \draw (5,0) -- node[below=5mm]{Potential ($V$)} (0,0) -- node[rotate=90,above]{Current ($I$)} (0,4);
                \footnotesize
                \draw [dashed,semithick] (2.5,3.5) -- (2.5,0);
                \draw [semithick] (2.5,0.1) -- ++(0,-0.2) node[below]{$E_{1/2}$};
    
                \draw [thick,blx,decoration={markings,mark=between positions 0.1 and 0.95 step 8mm with {\arrow{stealth}}},postaction={decorate}] (1,1.5) node[circle,fill,inner sep=1.5pt]{}
                    -- (1.8,1.5)
                    to[out=0,in=180] (3.2,3)
                    to[out=0,in=180] (4,2.5)
                    -- (3.5,2.5)
                    to[out=180,in=0] (1.8,1)
                    to[out=180,in=0] cycle
                ;
            \end{tikzpicture}
            \caption{Thermodynamics.}
            \label{fig:e-transferEnergya}
        \end{subfigure}
        \begin{subfigure}[b]{0.4\linewidth}
            \centering
            \begin{tikzpicture}[
                every node/.style={black}
            ]
                \small
                \draw (5,0) -- node[below=5mm]{Reaction Coordinate} (0,0) -- node[rotate=90,above]{Energy ($E$)} (0,4);
                \footnotesize
    
                \draw [thick,blx] (0.5,2.5)
                    -- node[below]{\ce{A^{n+1} + B^n}} (1.5,2.5)
                    to[out=0,in=180] (2.4,3.3)
                    to[out=0,in=180] (3.5,1.5)
                    -- node[below]{\ce{A^n + B^{n+1}}} (4.5,1.5)
                ;
    
                \begin{scope}[on background layer]
                    \draw [very thin] (1.5,2.5) -- (4.5,2.5);
                \end{scope}
                \draw [semithick,<->] (2.4,2.5) -- node[right,fill=white,inner sep=2pt]{$\Delta G^\ddagger$} (2.4,3.3);
                \draw [semithick,<->] (4,1.5) -- node[right]{$\Delta G$} (4,2.5);
            \end{tikzpicture}
            \caption{Kinetics.}
            \label{fig:e-transferEnergyb}
        \end{subfigure}
        \caption{Electron transfer reaction energies.}
        \label{fig:e-transferEnergy}
    \end{figure}
    \begin{enumerate}
        \item Thermodynamic:
        \begin{itemize}
            \item The difference in the potentials of \ce{A^{n+1}} and \ce{A^n}, and \ce{B^{n+1}} and \ce{B^n}.
            \begin{itemize}
                \item These can be measured electrochemically.
                \item We can measure the electrochemical driving force for these processes (i.e., the change in free energy during the reaction) with cyclic voltammetry.
            \end{itemize}
            \item In a cyclic voltammetry experiment\dots
            \begin{itemize}
                \item As we increase the potential to the point where the redox reaction will occur, we will see an increase as oxidation occurs.
                \item Then as we decrease the potential again to where the redox reaction will occur in the reverse direction, we will see a decrease as reduction occurs.
            \end{itemize}
            \item The midpoint $E_{1/2}$ is the thermodynamic potential (where redox is at equilibrium and you have equal amounts of both species). Is this $\Delta G$? What is going on here? Why are the equilibria misaligned?
        \end{itemize}
        \item Kinetics:
        \begin{itemize}
            \item $\Delta G=E_{1/2_\text{A}}-E_{1/2_\text{B}}$ where $E_{1/2_\text{X}}$ is the thermodynamic potential of substance X.
            \begin{itemize}
                \item $\Delta G$ is the thermodynamic contribution.
            \end{itemize}
            \item $\Delta G^\ddagger$ is the kinetic barrier, or activation energy.
        \end{itemize}
    \end{enumerate}
    \item The role of $\Delta G^\ddagger$ in an electron transfer.
    \begin{itemize}
        \item Electrons move very quickly and are highly delocalized with respect to the nuclei, so what dictates kinetics in these processes is nuclear motion (recall reorganization energy).
        \item In a simplistic sense, the key is the \ce{[A^n + B^{(n+1)}]^*} encounter complex.
        \item Electron transfer changes bond length.
        \begin{itemize}
            \item There is a kinetic barrier to the electron transfer because the thermodynamic energy is based on minimizing the energy in the reduced and oxidized forms.
        \end{itemize}
        \item Bond lengths change upon redox, so the solvent and countercations have to reorganize.
        \item This \textbf{reorganization energy} leads to a kinetic barrier (i.e., $\Delta G^\ddagger$).
        \item You can see evidence of the reorganization energy in Figure \ref{fig:e-transferEnergya}.
        \begin{itemize}
            \item You must go past the thermodynamic potential to observe the maximum/minimum current and attain complete oxidation/reduction.
        \end{itemize}
    \end{itemize}
    \item Measuring the reorganization energy.
    \begin{itemize}
        \item We do a self-exchange reaction with radiolabeled metal centers (see \textcite{bib:CHEM20100Notes}).
        \item Think of the energy scale on Figure IX.1 in \textcite{bib:CHEM20100Notes}) as discrete. To get over $\Delta G^\ddagger$, we must change vibrational states.
        \begin{itemize}
            \item Indeed, the short- and long-bond iron complexes have two vibrational states, but their combined transition state with medium bonds has a new vibrational state.
        \end{itemize}
        \item With electronic coupling, the two parabolas split into an upper loop and a lower loop with a bump.
        \item To treat this, we use the equation $\Delta G^\ddagger=\Delta G^\ddagger_t+\Delta G^\ddagger_v+\Delta G^\ddagger_0$.
        \begin{itemize}
            \item $\Delta G^\ddagger_t$ is the translational energy, which is moving the two species together.
            \item $\Delta G^\ddagger_v$ is vibrational, which is concerned with the bond lengths of the irons' matching structures.
            \item $\Delta G^\ddagger_0$ is the solvent, dipole, counterion, etc. This can be large (so one of the greatest contributors is the environment in which the system lies).
        \end{itemize}
    \end{itemize}
    \item Example:
    \begin{itemize}
        \item \ce{Co(NH3)6^2+} / \ce{Co(NH3)6^3+} is H.S. $d^7$ / L.S. $d^6$.
        \item Self-exchange is slow because nuclear reorganization is large ($\SI{0.2}{\angstrom}$ difference in bond length, which is significant).
        \begin{itemize}
            \item Note that this arises from the different electronic configurations.
        \end{itemize}
        \item The ion is getting smaller \emph{and} going low-spin during reorganization.
    \end{itemize}
    \item Another example:
    \begin{itemize}
        \item \ce{Ru(NH3)6^2+} / \ce{Ru(NH3)6^3+}.
        \item $k_\text{exch}$ is eight orders of magnitude faster than the previous example.
        \item This is because ruthenium is low-spin throughout ($\Delta(\ce{Ru-N})\approx\SI{0.04}{\angstrom}$ which is much smaller, so there is a smaller reorganization energy).
    \end{itemize}
    \item Key take aways:
    \begin{itemize}
        \item Electrons move fast, so what actually induces a kinetic barrier is the movement of the nuclei which have to reorganize in order to accommodate the electron popping between the two atoms.
        \item Can be accelerated by electron coupling, as in inner sphere mechanisms.
    \end{itemize}
    \item Inner sphere electron transfer: Some bonds are involved in the electron transfer.
    \begin{itemize}
        \item Accelerated by electron coupling, but hindered by greater nuclear reorganization energy (a bridging bond must be formed).
    \end{itemize}
    \item Example:
    \begin{itemize}
        \item Consider the reaction
        \begin{align*}
            \ce{Co(NH3)5Cl^2+ + Cr(H2O)5^2+} &\ce{->} \ce{Co(NH3)5^2+ + Cr(H2O)5Cl^2+}\\
            &\ce{->[H2O]} \ce{Co(H2O)6^2+ + Cr(H2O)5Cl^2+}
        \end{align*}
        \item The intermediates are \ce{[(H3N)5Co^{III}-Cl-Cr^{II}(OH2)5]^4+ -> [(H3N)5Co^{II}-Cl-Cr^{III}(OH2)5]^4+}.
        \item The rate is reasonably fast ($\SI{6e5}{\per\mole\per\second}$).
        \item How does this vary as a function of \ce{X-}?
        \begin{itemize}
            \item As ligand size (more diffuse; better at bridging) and charge (more electrostatic influences) increase, so does rate ($\ce{Br-}>\ce{Cl-}>\ce{F-}>\ce{H2O}>\ce{NH3}$).
        \end{itemize}
    \end{itemize}
    \item Inner-sphere electron transfer: Mixed valency.
    \item Consider the Creutz-Taube ion.
    \begin{figure}[H]
        \centering
        \chemleft{[}
            \chemfig{{(NH_3)_5}Ru-N**6(---N(-Ru{(NH_3)_5})---)}
        \chemright{]^{5+}}
        \caption{The Creutz-Taube ion.}
        \label{fig:CreutzTaube}
    \end{figure}
    \begin{itemize}
        \item The bridging ligand is a pyrazole.
        \item The electron transfer is very fast; thus, the oxidation state is approximately \ce{Ru2^{2.5}}.
    \end{itemize}
    \item Such electron transfers are measured with Near-IR spectroscopy, which can see inter-valence charge transfer bands (IVCT), which include bonds.
    \begin{itemize}
        \item This very low energy form of spectroscopy observes the energy that it takes to excite an electron between the two ruthenium centers.
    \end{itemize}
    \item Robin-Day classification:
    \begin{enumerate}[label={\Roman*)}]
        \item Completely localized.
        \begin{itemize}
            \item Regardless of the spectroscopic technique used, a difference between \ce{Ru^{II}} and \ce{Ru^{III}} can be observed.
        \end{itemize}
        \item Evidence of some delocalization.
        \begin{itemize}
            \item Most common.
        \end{itemize}
        \item Completely delocalized.
    \end{enumerate}
    \item If you go fast enough (ultrafast spectroscopy; femtosecond lasers), almost any system looks localized.
    \item Marcus theory:
    \begin{itemize}
        \item Built off of the Bell-Evans-Polanyi Relationship.
        \item See Figure IX.2 and the related discussion in \textcite{bib:CHEM20100Notes}.
        \item Two big insights:
        \begin{itemize}
            \item When $\Delta G=-\lambda$, $\Delta G^\ddagger=0$.
            \item When $\Delta G<-\lambda$, $\Delta G^\ddagger>0$.
        \end{itemize}
        \item The case where $\Delta G^\circ<-\lambda$ is called the \textbf{Marcus inverted region}.
        \item \textbf{Marcus equation}:
        \begin{equation*}
            k_\text{ET} = \nu_Nk_e\e[-\Delta G^\ddagger/RT]
        \end{equation*}
        where $\nu_N$ is the nuclear frequency (how accessible vibrational excited states are; related to the width of the parabolas in Figure IX.2 of \textcite{bib:CHEM20100Notes}) and $k_e$ is the electronic factor (related to overlap, probability of transfer, etc.; usually set to 1).
        \item More importantly, Marcus discovered that
        \begin{equation*}
            \Delta G^\ddagger = \frac{(\lambda+\Delta G)^2}{4\lambda}
        \end{equation*}
        which implies that as $\lambda\to -\Delta G$, $\Delta G^\ddagger\to 0$. Furthermore, as $\lambda$ passes $-\Delta G$, $\Delta G^\ddagger$ increases.
        \item This has important implications in biology, catalysis, etc. For example, if you want to slow down an undesirable side reaction and speed up your main reaction, provide more driving force. This accelerates your main reaction and moves your side reaction into the inverted region.
    \end{itemize}
\end{itemize}



\section{Lecture 6: Oxidative Addition and Reductive Elimination}
\begin{itemize}
    \item \marginnote{4/9:}Even further detached from organic chemistry (but scientists are looking for this reactivity in phosphorous and other $p$-block main group elements).
    \item General form:
    \begin{equation*}
        \ce{L_nM^Q + A-B <=>[{ox. adn.}][{red. elim.}] L_nM^{Q+2}AB}
    \end{equation*}
    \item Changes at the metal center during these two reactions (generically).
    \begin{itemize}
        \item Oxidative addition: Oxidation state, electron count, and coordination number increase by 2.
        \item Vice versa for reductive elimination.
    \end{itemize}
    \item Some notes on this reactivity:
    \begin{enumerate}
        \item Concerted reductive elimination must occur from a \emph{cis}-arrangement of ligands.
        \begin{itemize}
            \item Not true if it's stepwise.
        \end{itemize}
        \item Reductive elimination is favored by bulky ligands.
        \begin{itemize}
            \item Naturally: The more sterically crowded it is, the faster it will want to go.
        \end{itemize}
        \item Reductive elimination is disfavored for early metals.
        \begin{itemize}
            \item Since the early metals are very electropositive and don't want to access their lower oxidation states.
        \end{itemize}
        \item \ce{H-} as a reductive-elimination ligand is faster than other ligands.
        \begin{itemize}
            \item Due to it's spherically symmetric electron density.
            \item Hydrides kinetically (not thermodynamically) tend to react faster.
        \end{itemize}
        \item Oxidative addition can occur to give \emph{cis} or \emph{trans} products.
        \begin{itemize}
            \item The relative distribution of the stereochemistries gives us mechanistic information.
        \end{itemize}
    \end{enumerate}
    \item Classic studies: On Vaska's complex.
    \begin{figure}[h!]
        \centering
        \chemfig{Ir^{I}(-Cl)(-[1]PPh_3)(-[4]OC)(-[5]Ph_3P)}
        \caption{Vaska's complex.}
        \label{fig:vaskasComplex}
    \end{figure}
    \begin{itemize}
        \item Characteristics of the reactant: $d^8$, $16\,\e[-]$, canary yellow color.
        \item Characteristics of the product (after reacting with \ce{MeI}, \ce{H2}, or \ce{O2}): $d^6$, $18\,\e[-]$.
        \item \ce{MeI} bonds \emph{trans}, \ce{H2} bonds \emph{cis}, and \ce{O2} forms a peroxide.
        \item \ce{Ir^{III}} is inert, so we observe the kinetic products. These give us mechanistic information.
        \item Mechanistics can vary from S\textsubscript{N}2, to radical, to concerted, and so on.
        \begin{itemize}
            \item Classifying this reaction can get blurry. For instance, is \ce{M^Q + H^+ -> M^{Q+2}-H^+} an oxidative addition?
        \end{itemize}
    \end{itemize}
    \item Types of mechanisms:
    \begin{enumerate}
        \item Concerted.
        \begin{figure}[H]
            \centering
            \schemestart
                \chemfig{M}\arrow{0}[,0]\+{,,-1.25em}\chemfig{A-[6]B}
                \arrow
                \chemleft{[}
                    \chemfig{M?-[:30,,,,dashed]A-[6,,,,dashed]B?[,,dashed]}
                \chemright{]^\ddagger}
                \arrow
                \chemfig{M(-[:30]A)(-[:-30]B)}
            \schemestop
            \caption{The concerted mechanism for oxidative addition.}
            \label{fig:oxAdn-concerted-mechanism}
        \end{figure}
        \begin{figure}[H]
            \centering
            \begin{subfigure}[b]{0.35\linewidth}
                \centering
                \begin{tikzpicture}[scale=0.7]
                    \footnotesize
                    \fill [fill=blz] ({-6^0.5/4+0.2},0) circle (6^0.5/4);
                    \fill [fill=blz] ({6^0.5/4-0.2},0) circle (6^0.5/4);
                    \draw [semithick] ({-6^0.5/4+0.2},0) circle (6^0.5/4);
                    \draw [semithick] ({6^0.5/4-0.2},0) circle (6^0.5/4);
                    \node at (-1.3,0) {A};
                    \node at (1.3,0) {B};
                    \draw [semithick] (-2,0) circle (4mm);
                    \draw [semithick] (2,0) circle (4mm);
        
                    \begin{scope}[yshift=-3.6cm]
                        \node {M};
                        \draw [semithick,scale=1.5,xshift=-2mm] plot [domain=pi/2:3*pi/2,smooth,variable=\thta] ({\thta r}:{15^0.5/4*cos(\thta r)^2});
                        \draw [semithick,scale=1.5,xshift=2mm] plot [domain=-pi/2:pi/2,smooth,variable=\thta] ({\thta r}:{15^0.5/4*cos(\thta r)^2});
                        \filldraw [semithick,fill=blz,scale=1.5,rotate=90,xshift=-2mm] plot [domain=pi/2:3*pi/2,smooth,variable=\thta] ({\thta r}:{15^0.5/4*cos(\thta r)^2});
                        \filldraw [semithick,fill=blz,scale=1.5,rotate=90,xshift=2mm] plot [domain=-pi/2:pi/2,smooth,variable=\thta] ({\thta r}:{15^0.5/4*cos(\thta r)^2});
                    \end{scope}
        
                    \node [rotate=-90] at (0,-1.2) {\normalsize$:\longrightarrow$};
                \end{tikzpicture}
                \caption{HOMO interaction.}
                \label{fig:oxAdn-concerted-orbitalsa}
            \end{subfigure}
            \begin{subfigure}[b]{0.35\linewidth}
                \centering
                \begin{tikzpicture}[scale=0.7]
                    \footnotesize
                    \filldraw [semithick,fill=blz] ({-6^0.5/4-1.6},0) circle (6^0.5/4);
                    \filldraw [semithick,fill=white] ({6^0.5/4+1.6},0) circle (6^0.5/4);
                    \node at (-1.3,0) {A};
                    \node at (1.3,0) {B};
                    \filldraw [semithick,fill=white] (-0.6,0) circle (4mm);
                    \filldraw [semithick,fill=blz] (0.6,0) circle (4mm);
        
                    \begin{scope}[yshift=-3.6cm]
                        \node {M};
                        \filldraw [semithick,fill=white,scale=2,xshift=-1.5mm,yshift=-1.5mm] plot [domain=-pi:-pi/2,smooth,variable=\thta] ({\thta r}:{15^0.5/2*cos(\thta r)*sin(\thta r)});
                        \filldraw [semithick,fill=blz,scale=2,xshift=-1.5mm,yshift=1.5mm] plot [domain=-pi/2:0,smooth,variable=\thta] ({\thta r}:{15^0.5/2*cos(\thta r)*sin(\thta r)});
                        \filldraw [semithick,fill=white,scale=2,xshift=1.5mm,yshift=1.5mm] plot [domain=0:pi/2,smooth,variable=\thta] ({\thta r}:{15^0.5/2*cos(\thta r)*sin(\thta r)});
                        \filldraw [semithick,fill=blz,scale=2,xshift=1.5mm,yshift=-1.5mm] plot [domain=pi/2:pi,smooth,variable=\thta] ({\thta r}:{15^0.5/2*cos(\thta r)*sin(\thta r)});
                    \end{scope}
        
                    \node [rotate=110] at (-1.8,-1.2) {\normalsize$\cdot\longrightarrow$};
                    \node [rotate=70] at (1.8,-1.2) {\normalsize$\cdot\longrightarrow$};
                \end{tikzpicture}
                \caption{LUMO interaction.}
                \label{fig:oxAdn-concerted-orbitalsb}
            \end{subfigure}
            \caption{Orbital interactions in concerted oxidative addition.}
            \label{fig:oxAdn-concerted-orbitals}
        \end{figure}
        \begin{itemize}
            \item \emph{cis} products.
            \begin{itemize}
                \item If you see a \emph{trans} product, this is not the mechanism.
            \end{itemize}
            \item HOMO interaction: The \ce{A-B} $\sigma$-bonding orbital donates to an empty metal $d_{x^2-y^2}$-symmetry orbital.
            \item LUMO interaction: The metal $d_{xy}$ orbital backbonds into the \ce{A-B} $\sigma^*$ antibonding orbital. It is these electrons that depopulate the $\sigma$ bond and allow the \ce{A-B} bond to split.
        \end{itemize}
        \item S\textsubscript{N}2.
        \begin{equation*}
            \ce{\charge{0=\:}{M}\; + R-X -> M^+-R + \charge{90=\:}{X}^- -> X-M-R}
        \end{equation*}
        \begin{itemize}
            \item \emph{cis} or \emph{trans}; no clear preference.
            \item One signature of this reactivity is a steric preference where $\text{primary}>\text{secondary}>\text{tertiary}$.
            \begin{itemize}
                \item Some of the trends of organic reactions can appear in inorganic reactions!
            \end{itemize}
        \end{itemize}
        \item Radical chain.
        \begin{align*}
            \ce{M + In} &\ce{->} \ce{MIn*}\\
            &\ce{->[R-X]} \ce{InM-X + R*}\\
            &\ce{->[M]} \ce{R-M*}\\
            &\ce{->[R-X]} \ce{R-M-X + R*}
        \end{align*}
        \begin{itemize}
            \item We begin with a metal and an initiator (denoted In).
            \begin{itemize}
                \item These react to form the adduct \ce{MIn*}, which is a radical.
                \item This radical grabs an \ce{X} from \ce{R-X} to form \ce{InM-X} (this product is worthless and the final step of initiation). A \ce{R*} is also generated in this step (this is the radical to be propagated).
                \item The \ce{R*} reacts with another equivalent of the metal to create an \ce{R-M*} radical which can then grab a halide, generating an \ce{R-M-X} and an \ce{R*}, the former of which is the desired product and the latter of which can continue propagating.
            \end{itemize}
            \item Pretty common.
        \end{itemize}
        \item Electron transfer.
        \begin{align*}
            \ce{M^0 + R-X} &\ce{->} \ce{[M^+* + RX^-*]}\\
            &\ce{->} \ce{M^+* + R* + X^-}\\
            &\ce{->} \ce{M^+-R + X^-}\\
            &\ce{->} \ce{X-M-R}
        \end{align*}
        \begin{itemize}
            \item Related to S\textsubscript{N}2 again.
            \item This is not a radical chain process (no propagation); it's just an electron transfer followed by fragmentation and recombination.
        \end{itemize}
    \end{enumerate}
    \item Jack Halpern of UChicago writes \textcite{bib:HalpernVaska}.
    \begin{itemize}
        \item Studies \ce{L2IrClCO ->[MeX] L2IrClCOMeX}, specifically the reaction rates as a function of \ce{X}.
        \item Determines that $\ce{I-}>\ce{Br-}>>\ce{Cl-}$, and that $\text{Rate}=k\ce{[CH3X][Ir]}$.
        \item $\Delta S^\ddagger=\SI{-43}{{e.u.}}$ (where an e.u. is an entropy unit).
        \item Solvent effects are also consistent with an ionic mechanism.
        \item There is a 5-coordinate intermediate, and the two new ligands end up adding \emph{trans}.
        \item If you use \ce{EtI}, this reaction proceeds through a radical mechanism.
        \begin{itemize}
            \item You have mechanistic switching based on the identity of the substrate.
        \end{itemize}
    \end{itemize}
    \item Bimolecular oxidative addition:
    \begin{equation*}
        \ce{2L_nM^0 + R-X -> L_nM^I-R + L_nM^I-X}
    \end{equation*}
    \item Jack Halpern again writes \textcite{bib:HalpernBimolecularOxAdn}.
    \begin{itemize}
        \item Studies the reaction \ce{2Co^{II}(CN)5^3- + MeI -> Co^{III}(CN)5Me^3- + Co^{III}(CN)5I^3-}.
        \item Follows some transition metal trends:
        \begin{itemize}
            \item Heavier metals (such as iridium above) do 2-electron chemistry.
            \item Lighter metals (such as this one) do 1-electron chemistry.
        \end{itemize}
        \item We can also see radical-type reactivity.
    \end{itemize}
    \item Another parallel example: \ce{2L_nM + 2RX -> 2L_nM-X + R-R}.
    \begin{itemize}
        \item For example, \ce{Cp^*2Yb(OEt2) + CH2Cl2 -> 2Cp^*2Yb-Cl + ClCH2-CH2Cl}.
        \item This can also occur through a radical mechanism. We form an \ce{R*} radical that can recombine to form \ce{R-R} or make \ce{M-R}.
    \end{itemize}
    \item Stereochemistry of S\textsubscript{N}2 type reactivity.
    \item George Whitesides writes \textcite{bib:WhitesidesHDFpInversion}.
    \begin{figure}[H]
        \centering
        \begin{subfigure}[b]{0.3\linewidth}
            \centering
            \chemfig{(<[:-120]H)(>:[:-160]D)(-[:120]{\emph{t}-}Bu)-(>:[:60]D)(<[:20]H)-[:-60]OTs}
            \vspace{2em}
            \caption{The reactant.}
            \label{fig:SN2-stereochemistrya}
        \end{subfigure}
        \begin{subfigure}[b]{0.6\linewidth}
            \centering
            \begin{tikzpicture}[
                every node/.style={circle}
            ]
                \begin{scope}
                    \draw circle (0.5cm);
                    \draw 
                        (0,0) -- (90:0.8) node[above=-2mm]{$t$-Bu}
                        (30:0.5) -- (30:0.8) node[anchor=-150]{H}
                        (0,0) -- (-30:0.8) node[anchor=150]{H}
                        (-90:0.5) -- (-90:0.8) node[anchor=113,yshift=2mm]{OTs}
                        (0,0) -- (-150:0.8) node[anchor=30]{D}
                        (150:0.5) -- (150:0.8) node[anchor=-30]{D}
                    ;
                \end{scope}
                \draw [semithick,-stealth] (1.5,0) -- node[above=-2mm]{\small\ce{Fp-}} (2.5,0);
                \begin{scope}[xshift=4cm]
                    \draw circle (0.5cm);
                    \draw 
                        (0,0) -- (90:0.8) node[above=-2mm]{$t$-Bu}
                        (30:0.5) -- (30:0.8) node[anchor=-150]{D}
                        (0,0) -- (-30:0.8) node[anchor=150]{H}
                        (-90:0.5) -- (-90:0.8) node[below=-1mm]{Fp}
                        (0,0) -- (-150:0.8) node[anchor=30]{D}
                        (150:0.5) -- (150:0.8) node[anchor=-30]{H}
                    ;
                \end{scope}
            \end{tikzpicture}
            \caption{The reaction.}
            \label{fig:SN2-stereochemistryb}
        \end{subfigure}
        \caption{S\textsubscript{N}2 stereochemistry.}
        \label{fig:SN2-stereochemistry}
    \end{figure}
    \begin{itemize}
        \item Observes inversion (by looking at $\text{J}_{\ce{H-H}}$ coupling by NMR) of \ce{H} and \ce{D} at a single stereocenter.
        \item React the compound in Figure \ref{fig:SN2-stereochemistrya} with a \ce{Fp-} fragment.
        \item Observe inversion, as in Figure \ref{fig:SN2-stereochemistryb}, so it's S\textsubscript{N}2.
    \end{itemize}
    \item Another synthesis:
    \begin{figure}[h!]
        \centering
        \schemestart
            \chemfig{(<[:-120]D)(>:[:-160]H)(-[:120]TsO)-(>:[:60]H)(<[:20]D)-[:-60]{\emph{t}-}Bu}
            \arrow{->[\small\ce{Pd^0}]}
            \chemfig{(<[:-120]H)(>:[:-160]D)(-[:120]Pd-[1]OTs)-(>:[:60]H)(<[:20]D)-[:-60]{\emph{t}-}Bu}
            \arrow{->[\small\ce{PhMgBr}]}[,1.2]
            \chemfig{(<[:-120]H)(>:[:-160]D)(-[:120]Pd-[1]Ph)-(>:[:60]H)(<[:20]D)-[:-60]{\emph{t}-}Bu}
            \arrow{->[][\footnotesize{red. elim.}]}[,1.2]
            \chemfig{(<[:-120]H)(>:[:-160]D)(-[:120]Ph)-(>:[:60]H)(<[:20]D)-[:-60]{\emph{t}-}Bu}
        \schemestop
        \caption{An additional way of probing S\textsubscript{N}2 addition/elimination.}
        \label{fig:SN2-additionalSynthesis}
    \end{figure}
    \begin{itemize}
        \item This one shows us that palladium causes an inversion once again, but reductive elimination does not.
    \end{itemize}
    \item Radical mechanisms.
    \begin{itemize}
        \item We probe these with \textbf{radical clocks}.
        \item The unzipping of the methylcyclopropane radical ring happens so fast that it will necessarily be faster than any recombination with \ce{M-Cl}.
        \item Since iodides are easier to reduce than bromides, bromomethylcyclopropane will react in a straightforward manner with \ce{Fp-}, but iodomethylcyclopropane and \ce{Fp-} will pursue a radical mechanism to a competitive degree. Why does the reduction potential of iodides and bromides matter?
    \end{itemize}
    \item \textbf{Radical clock}: A reagent such that if a radical is generated on it, it will undergo a rapid isomerization or redistribution to generate different product(s).
    \item A few notes.
    \begin{enumerate}
        \item Similar rules to those in orgo apply.
        \begin{itemize}
            \item For example, \ce{I-} is a better leaving group than \ce{Cl-}.
            \item However, there are exceptions, too: \ce{CN} is a terrible leaving group in inorganic chemistry, whereas you can sometimes kick it out in orgo.
        \end{itemize}
        \item Sterics matter.
        \begin{itemize}
            \item Oxidative addition is slower for sterically encumbered substrates.
            \item If you want to favor a radical reaction over a concerted or nucleophilic mechanism, make the compound bulky. This will disfavor the two undesired mechanisms but not an electron transfer step.
        \end{itemize}
        \item First row metals will be faster than second and third row metals.
        \begin{itemize}
            \item This is because they're much more reactive. However, they will also go down competitive side paths more readily (can be good or bad).
            \item Because of this, second and third row metals are more often used. Plus, you can just heat them up a bit to speed up the reaction.
        \end{itemize}
    \end{enumerate}
\end{itemize}



\section{Lecture 7: Insertion/Deinsertion and Kinetics}
\begin{itemize}
    \item \marginnote{4/12:}Migratory insertion/deinsertion.
    \item Also pretty unique to the transition metals.
    \item General form:
    \begin{equation*}
        \ce{L-M^Q-X <=>[{insertion}][{deinsertion}] M^Q-L-X}
    \end{equation*}
    \begin{itemize}
        \item In the course of this reaction, the \ce{L} is converted into an X-type ligand.
    \end{itemize}
    \item Characteristics of insertion: Electron count decreases by 2, coordination number decreases by 1, and the oxidation state does not change.
    \item Examples:
    \begin{itemize}
        \item 1,1-insertion: \ce{Me-M-CO <=> M-C(=O)-Me}.
        \begin{itemize}
            \item So named because the metal and the migrating group end up at the same position on the carbonyl ligand (the 1 position).
        \end{itemize}
        \item 1,2-migration: \ce{Cp2Zr(-||)Me <=> Cp2ZrPr}.
        \begin{itemize}
            \item So named because the metal ends up on the 1 position of the ethylene olefin and the migrating group ends up on the 2 position of the ethylene olefin (remember that we number substituents from the metal center outwards).
        \end{itemize}
    \end{itemize}
    \item More groups than methyl can migrate; it's just that methyl commonly migrates.
    \item Insertions into \ce{M-C} bonds are common.
    \begin{itemize}
        \item Insertions into \ce{M-H} bonds are common for olefins, but uncommon for \ce{CO} because metal formyl species are unstable.
        \item You can also insert into \ce{M-O} bonds (note that dppe stands for diphenylphosphinoethane):
        \begin{figure}[h!]
            \centering
            \schemestart
                \chemfig{Pt?(-[3]\chemabove{P}{Ph_2}-[:-150]-[6,0.9]-[:-30]\chembelow{P}{Ph_2}?)(-[7]Me)(-[1]OMe)}
                \arrow(.east--.-173){->[\small\ce{CO}]}
                \chemfig{{(dppe)}Pt(-[7]Me)(-[:30](=[2,0.7]O)-[:-30]OMe)}
            \schemestop
            \vspace{1em}
            \caption{Insertion into an \ce{M-O} bond.}
            \label{fig:insertion-M-O}
        \end{figure}
    \end{itemize}
    \item A note on the mechanism:
    \begin{itemize}
        \item We can either take the perspective that the X group migrates or that the L group inserts itself into the \ce{M-X} bond.
        \item Thus, either the $\sigma$ bond of the migrating ligand attacks the site to which it bonds or the L group moves into the $\sigma$ bond.
        \item We call this a migratory insertion, but there are two possible mechanisms (it's hard to know what is migrating and what is staying put).
        \item Answer: The X-type ligand is migrating. We can test this by radiolabeling one of the carbonyls in \ce{Mn(CO)5Me}?
    \end{itemize}
    \item $\beta$-\ce{H} elimination: \ce{L2NiEt <=> L2Ni(-||)H}.
    \begin{figure}[H]
        \centering
        \chemleft{[}
            \chemfig{\charge{-30=\:}{Ni}?-[1]-[:-30]-[:-110]H?[,,dashed]}
        \chemright{]^\ddagger}
        \caption{The transition state in a $\beta$-\ce{H} elimination.}
        \label{fig:betaHydrideTransState}
    \end{figure}
    \begin{itemize}
        \item The transition state (see Figure \ref{fig:betaHydrideTransState}) shows an \textbf{agostic interaction}.
    \end{itemize}
    \item $\alpha$-elimination\footnote{Note that Ar stands for an aryl group.}.
    \begin{figure}[H]
        \centering
        \schemestart
            \chemfig{Pt?(-[3]\chemabove{P}{R_2}-[:-150]-[6,0.9]-[:-30]\chembelow{P}{R_2}?)(-[1]SiAr_2(-[:120,0.7,1]H))(-[7]Me)}
            \+
            \chemfig{\ce{B(C6F5)3}}
            \arrow(.-7--.west){->[][\small\ce{-MeB(ArF)3^-}]}[0,1.7]
            \chemleft{[}
                \chemfig{\charge{120[yshift=1mm]=$\oplus$}{Pt}?-[1]\chemabove{Si}{Ar_2}-[7]H?[,,dashed]}
            \chemright{]^\ddagger}
            \arrow
            \chemleft{[}
                \chemfig{Pt?(-[3]\chemabove{P}{R_2}-[:-150]-[6,0.9]-[:-30]\chembelow{P}{R_2}?)(-[7]H)(=[1]SiAr_2)}
            \chemright{]^+}
        \schemestop
        \caption{An example of $\alpha$-elimination.}
        \label{fig:alphaElimination}
    \end{figure}
    \item External attack at a ligand.
    \begin{figure}[H]
        \centering
        \ce{L_nM^Q=X + Nu- <=> [L_nM^{Q-2}-X-Nu]-}\\[1em]
        \schemestart
            \chemfig{\ce{L_n}M^{\emph{Q}}-\phantom{i}-[6,0.3]Y=[2]X}
            \arrow{->[\small\ce{Nu-}]}
            \chemleft{[}
                \chemfig{\ce{L_n}M^{\emph{Q}}-[:30]X-[:-30]Y-[:30]Nu}
            \chemright{]^-}
        \schemestop\\[1em]
        \schemestart
            \chemfig{\ce{L_n}M^{\emph{Q}}-X~Y}
            \arrow{->[\small\ce{Nu-}]}
            \chemleft{[}
                \chemfig{\ce{L_n}M^{\emph{Q}}-[:30]X(=[2]Y)-[:-30]Nu}
            \chemright{]^-}
        \schemestop
        \caption{Types of external attack at a ligand.}
        \label{fig:externalLigandAttack}
    \end{figure}
    \begin{itemize}
        \item Somewhat more related to organic chemistry.
        \item Lists some examples.
    \end{itemize}
    \item \ce{Tp} is trispyrazolylborate.
    \begin{itemize}
        \item It's a \ce{Cp} analogue, meaning that it has the same electron count and similar sterics.
    \end{itemize}
    \item Be aware of Fischer carbenes.
    \item There could be a radical process.
    \begin{itemize}
        \item \textcite{bib:CrevierMayer} tells us that an osmium-nitrido external attack at a ligand must be a 2-electron process, not a radical mechanism.
    \end{itemize}
    \item Electrophilic attack on a ligand.
    \item Example: \ce{Ir^{II}(PPh3)2HCl(NO) ->[HCl] Ir^{III}(PPh3)2HCl2(N(=O)H)}.
    \begin{itemize}
        \item The reactant is a $16\,\e[-]$ species.
        \item The nitrogen-containing ligand is a nitroxyl ligand.
    \end{itemize}
    \item Gives some other examples.
    \item \ce{Tp^*} is a \ce{Tp} group where each pyrazole is 3,5-dimethyl substituted.
    \item Sometimes we create a positive metal cation. This can be accomplished either via a direct electrophilic attack on an attached R group or via an attack at the metal followed by reductive elimination.
    \item $\sigma$-bond metathesis: \ce{L_nM^Q-X + Y-Z <=> L_nM^Q-Y + X-Z}.
    \begin{itemize}
        \item Usually observed for $d^0$ systems.
    \end{itemize}
    \item Example: \ce{Zr^{IV}(N(SiR3)H)3Me -> Zr(N(SiR3)H)2(=N-SiR3) + CH4}.
    \begin{itemize}
        \item \ce{C-H} activation is a big thing in synthetic chemistry, and a lot of the pathways go through $d^0$, early, reactive transition metals.
    \end{itemize}
    \item More on $\sigma$-bond metathesis:
    \begin{enumerate}
        \item Most common for early metals.
        \begin{itemize}
            \item Especially $d^0$ metals.
        \end{itemize}
        \item Thought to go through a 4-membered transition state.
        \item There is likely a continuum between "pure" $\sigma$-bond metathesis and oxidative addition/reductive elimination.
        \item This still requires an open coordination site and $\leq 16\,\e[-]$.
        \begin{itemize}
            \item Because the first step is coordination, usually to form some kind of $\sigma$-adduct.
        \end{itemize}
    \end{enumerate}
    \item Kinetics of associative substitution.
    \begin{equation*}
        \text{rate} = -\dv{\ce{[ML_x]}}{t}
        = \dv{\ce{[ML_{x-1}L$'$]}}{t}
        = k\ce{[ML_x][L$'$]}
    \end{equation*}
    \begin{figure}[h!]
        \centering
        \begin{subfigure}[b]{0.45\linewidth}
            \centering
            \begin{tikzpicture}[
                every node/.style={black}
            ]
                \small
                \draw (5,0) -- (0,0) -- node[left]{$\ln\ce{[ML_x]}$} (0,4);
                \footnotesize
                \draw (2,0.1) -- ++(0,-0.2) node[below]{$t_{1/2}$};
    
                \draw [blx,thick] (0.2,3.7) -- (4.3,0.5);
                \draw [grx,thick] (0.2,3.5) -- (1.9,0.2) node[above right,yshift=-1mm]{high \ce{[L$'$]}};
                \draw [rex,thick] (0.35,3.7) -- (4.3,2.5) node[below]{Low \ce{[L$'$]}};
            \end{tikzpicture}
            \caption{Reaction coordinate and \ce{[ML_x]}.}
            \label{fig:associativeKineticsa}
        \end{subfigure}
        \begin{subfigure}[b]{0.45\linewidth}
            \centering
            \begin{tikzpicture}[
                every node/.style={black}
            ]
                \small
                \draw (5,0) -- node[below]{\ce{[L$'$]}} (0,0) -- node[left]{$k_\text{obs}$} (0,4);
                \footnotesize
    
                \draw [blx,thick] (0.2,0.2) -- (4,3.4) node[below right,yshift=1mm]{Theory};
                \draw [grx,thick] (0.2,0.2) to[out=52,in=-140] (4,3.8) node[above right,yshift=-1mm]{Experiment};
            \end{tikzpicture}
            \caption{Rate and \ce{[L$'$]}.}
            \label{fig:associativeKineticsb}
        \end{subfigure}
        \caption{Kinetics of associative substitution.}
        \label{fig:associativeKinetics}
    \end{figure}
    \begin{itemize}
        \item Experimentally, we can use a large \ce{[L$'$]} to get to pseudo-first order conditions.
        \item This gives us $\text{rate}=k_\text{obs}\ce{[ML_x]}$ where $k_\text{obs}=k\ce{[L$'$]}$.
        \item In Figure \ref{fig:associativeKineticsa}, $t_{1/2}$ is a midpoint, and the rate gets faster (steeper slope) with more \ce{[L$'$]} and slower (more gradual slope) with less \ce{[L$'$]}.
        \item There is a discrepancy between theory and experiment in Figure \ref{fig:associativeKineticsb}.
        \begin{itemize}
            \item This is because of the presence of a solvent-assisted mechanism.
            \item \ce{L_xM + {solv} <=>[$k_s$] L_{x-1}ML({solv}) ->[L$'$][{fast}] L_{x-1}ML$'$ + {solv} + L}.
            \item Note that $k_s$ is the rate of solvent association.
        \end{itemize}
        \item The solvent-assisted mechanism dominates at low \ce{[L$'$]}, and vice versa for the normal mechanism.
    \end{itemize}
    \item Kinetics of dissociative substitutions.
    \begin{enumerate}[label={\roman*)}]
        \item \ce{ML_x <=>[$k_1$][$k_{-1}$] [ML_{x-1}] + L}.
        \item \ce{[ML_{x-1}] + L$'$ ->[$k_2$] ML_{x-1}L$'$} (assume irreversible).
    \end{enumerate}
    \begin{itemize}
        \item There are now two cases:
        \begin{enumerate}[label={\alph*)}]
            \item Fast pre-equilibrium, i.e., $k_1,k_{-1}>>k_2$. This gives us
            \begin{equation*}
                \text{rate} = k_1k_2\cdot\frac{\ce{[ML_x][L$'$]}}{\ce{[L]}}
            \end{equation*}
            \item Steady state approximation: $
                \dv*{\ce{[ML_{x-1}]}}{t} = 0
                = k_1\ce{[ML_x]}-k_{-1}\ce{[ML_{x-1}]}-k_2\ce{[ML_{x-1}][L$'$]}
            $. If we solve the above for \ce{[ML_{x-1}]}, then we get
            \begin{equation*}
                \text{rate} = \frac{k_1k_2\ce{[ML_x][L$'$]}}{k_{-1}\ce{[L]}+k_2\ce{[L$'$]}}
            \end{equation*}
            If we now assume that $k_2\ce{[L$'$]}$ is large, then we get $\text{rate}=k_1\ce{[ML_x]}$.
            \begin{itemize}
                \item The steady state approximation is a good assumption to make because if you see a buildup of the dissociative intermediate, you can measure the rates. Alternatively, if you don't see it, you can assume that $\ce{[ML_{x-1}]}=0$.
            \end{itemize}
        \end{enumerate}
        \item To do this experimentally, we add a large concentration of \ce{[L$'$]}, or \ce{[L]} in some cases.
        \begin{itemize}
            \item This gives us $\text{rate}=k_\text{obs}\ce{[ML_x]}$ where $k_\text{obs}$ denotes the mess from the above equation.
        \end{itemize}
        \begin{figure}[h!]
            \centering
            \begin{tikzpicture}
                \small
                \draw (5,0) -- node[below]{\ce{[L$'$]}} (0,0) -- node[left]{$k_\text{obs}$} (0,4);
                \footnotesize
        
                \draw [semithick,dashed] (0,2.5) -- (4.5,2.5);
                \draw [blx,thick] (0,0) to[out=60,in=-175] (4.5,2.3);
        
                \node [anchor=south west] at (1.3,0.1) {2nd order region}
                    edge [->,out=180,in=-60] (0.7,0.8)
                ;
                \node [anchor=south east] at (3.5,2.6) {1st order region}
                    edge [->,out=0,in=120] (4,2.35)
                ;
            \end{tikzpicture}
            \caption{The effect of \ce{[L$'$]} on rate.}
            \label{fig:LPrimeRate}
        \end{figure}
        \item If we plot $k_\text{obs}$ vs. \ce{[L$'$]}, we get Figure \ref{fig:LPrimeRate}.
        \begin{itemize}
            \item In the second order region, \ce{ML_x <=>[{fast}] ML_{x-1}} many times before product formation.
            \item In the first order region, every time \ce{ML_{x-1}} forms, it goes on to become a product.
        \end{itemize}
        \item In other words, $k_\text{obs}$ should approach $k_1$ when $k_2\ce{[L$'$]}>>k_{-1}\ce{[L]}$.
        \item Importantly, as \ce{L$'$} is varied with different ligands, $k_1$ should stay constant (assuming the mechanism doesn't change).
        \item A double reciprocal plot can be used to obtain still more information about the reaction.
        \begin{itemize}
            \item If you plot $\frac{1}{k_\text{obs}}$ vs. $\frac{\ce{[L]}}{\ce{[L$'$]}}$ and run a linear regression, the slope will be $\frac{k_{-1}}{k_1k_2}$ and the $y$-intercept will be $\frac{1}{k_1}$.
        \end{itemize}
    \end{itemize}
    \item Transition state theory basics:
    \begin{itemize}
        \item In 1887, Arrhenius comes up with the Arrhenius equation $k=A\e[-E_A/RT]$, which can be algebraically manipulated into
        \begin{equation*}
            \ln k = \ln A-\frac{E_A}{RT}
        \end{equation*}
        where $E_A$ is the activation energy, $R$ is the gas constant, and $T$ is temperature.
        \begin{itemize}
            \item This allows us to create a linear $\ln k$ vs. $\frac{1}{T}$ plot from which we can pull out important information.
        \end{itemize}
        \item In the 1930s, Eyring comes up with the Eyring equation
        \begin{equation*}
            \ln\left( \frac{k}{T} \right) = \frac{-\Delta H^\ddagger}{RT}+\ln\left( \frac{k_B}{h} \right)+\frac{\Delta S^\ddagger}{R}
        \end{equation*}
        where we can calculate that $\ln(k_B/h)\approx 23.76$.
        \begin{itemize}
            \item This allows us to create a linear $\ln(k/T)$ vs. $\frac{1}{T}$ plot from which we can pull out additional important information.
        \end{itemize}
    \end{itemize}
\end{itemize}



\section{Office Hours (Anderson)}
\begin{itemize}
    \item Explain cyclic voltammetry.
    \begin{itemize}
        \item SOP for CV in the modules explains cyclic voltammetry.
        \item You have a working and a reference electrode, as well as an auxiliary platinum wire. All of these are submerged in the same solution (of a polar solvent and your compound).
        \item You apply a voltage across the working and auxiliary electrodes.
        \item With nothing in your solution, you have pseudo-capacitance/polar something-or-other. It's basically just a loop with spikes on the end ($D_{2h}$ symmetry) in your voltage vs. current graph.
        \item However, if we have a compound, we pass some current in the forward and reverse directions at the thermodynamic potential point (the free energy of the reduction/oxidation vs. some other reduction/oxidation process; we can convert to $\Delta G$ with the Nernst equation); the spikes aren't entirely superimposable because we must push the equilibrium a bit past to get something to happen.
    \end{itemize}
    \item How do you handle ligands with charges? Like \ce{SO4^2-} in question 8 of the last homework or the allyl group in 1.16. Where do the electrons come from?
    \begin{itemize}
        \item 1.16:
        \begin{itemize}
            \item Redraw as a legitimate Lewis structure.
            \begin{center}
                \chemleft{[}
                    \chemfig{Pd?(-[:150]-[:-110]=^[:-60]-[:120,0.3,,,white]\phantom{i}?[,,{stealth-}])(-[1]PCy_3)(-[7]OEt_2)}
                \chemright{]^+}
            \end{center}
            \item Allyl is 1 covalent, 1 dative; the other two ligands are each dative donors.
            \item Thus, palladium's oxidation state is \ce{Pd^2+}.
            \item The synthetic route by which we get to this compound has no bearing on it's properties (electron count, oxidation state, etc.).
            \item In one reaction mechanism, it's \ce{C3H5+} before, and then it's \ce{C3H5-} after splitting.
        \end{itemize}
        \item 8:
        \begin{itemize}
            \item 8b: Solve with electron counts? \ce{MoL2X4} is $14\,\e[-]$, whereas \ce{MoL4X2} is $16\,\e[-]$.
            \item 8d: Donates the same number of electrons in each case! They're just different resonance structures!!
            \item 8e: Cp is an \ce{L2X}-type ligand with $\text{L.B.N.}=3$, for example.
            \item 8f: Ligand domains are useful, but not every inorganic chemist uses them.
        \end{itemize}
    \end{itemize}
    \item Intuition for reductive elimination?
    \begin{itemize}
        \item Yes, it is the reverse of oxidative addition.
        \item Generally, you break the \ce{M-B} bond and then the \ce{M-A} bond.
        \item You could also break \ce{M-B} and \ce{A-B}, and then have \ce{B-} nucleophilically attack \ce{[M-A]+}.
        \item You could also do this radical-wise: \ce{M-A$\cdot$ + $\cdot$B -> M + A-B}.
        \item Could be thought of as concerted, but probably best not to.
    \end{itemize}
    \item To do one-electron chemistry, you have to stabilize radicals.
    \begin{itemize}
        \item You have more one-electron chemistry in high-spin first-row transition metals.
    \end{itemize}
    \item There is a synthesis question on the Homework 2, so start looking at that sooner as opposed to later.
\end{itemize}



\section{Office Hours (Whitmeyer)}
\begin{itemize}
    \item \marginnote{4/13:}The fic fragment: Right spelling/charge?
    \item Does the iron in Figure \ref{fig:FpFragments} have tetrahedral or square planar geometry and why?
    \item Copies Table VI.2 from \textcite{bib:CHEM20100Notes}. Some differences?
    \item Why these specific spins to higher and lower energy (Figure \ref{fig:zeemanSplitting-1e})? Or is it arbitrary?
    \begin{itemize}
        \item Convention that negative is lower.
    \end{itemize}
    \item What is a Boltzmann population?
    \begin{itemize}
        \item A Boltzmann distribution is a mathematical distribution much like the normal distribution.
        \item There is an ideal bond distance, and some atoms bond closer or farther.
    \end{itemize}
    \item Differences in Curie's Law from last quarter?
    \item In this course, we include units for equilibrium constants, derived from the mass-action expression.
    \item HW2 4a:
    \begin{itemize}
        \item First row compounds are less likely to delocalize?
        \item LFSE is larger for second, third row.
        \item First row complexes are more likely to have higher spin states.
        \item Bigger orbitals with more overlap are better at electron transfer.
    \end{itemize}
\end{itemize}



\section{Discussion Section}
\begin{itemize}
    \item HW2 due 4/19 at 12 PM CT.
    \item Midterm 1: 4/22 from 6pm-8pm CT (proctored by Dr. Anderson).
    \item No discussion section or office hour on 4/20; there will however be a 2 hour exam review and office hour session on 4/17 from 11:30-1:30 CT.
    \item Sophie will not be available at all Midterm week.
    \item I can send discussion topics ahead of time or post in the Discussion on Canvas.
    \item More information will be forthcoming on where Sophie took off points on our homework.
    \item Zirconium is quasi-stable at $16\,\e[-]$.
\end{itemize}



\section{Lecture 8: Alkyls, Aryls, and Multiple Bonds}
\begin{itemize}
    \item \marginnote{4/14:}Done with reactions; we're now onto ligand types.
    \item Alkyls and aryls are the quintessential organometallic ligands.
    \item \textbf{Alkyl}: An \ce{M-CR3} ligand where the carbon is $sp^3$ hybridized.
    \item Historically, the stability of these compounds has hindered their isolation.
    \item The first alkyl compound to be characterized was \ce{ZnEt2}, synthesized in 1847 by E. Franklin at Imperial College.
    \begin{itemize}
        \item He reacted \ce{Zn^0 + EtI -> ZnEt2(l)}.
        \item \ce{ZnEt2} is extremely flammable, can be distilled, and is a good alkylating agent.
    \end{itemize}
    \item 50 years later, Victor Grignard reacts
    \begin{equation*}
        \ce{Mg^0 + R-X -> XMgR({solv})2 <=> MgR2 + MgX2 ->[{dioxane}] X2Mg-{diox}-MgX2}
    \end{equation*}
    \begin{itemize}
        \item The equilibrium between the solvate and \ce{MgR2 + MgX2} is called the \textbf{Schlenk equilibrium}. It can be pushed one way or the other with certain reagents.
        \item For example, adding dioxane precipitates out the magnesium halides, leaving you with only the desired alkyl halide Grignard reagent in solution.
        \item Note that dioxane is a six-membered single-bonded ring with para-oxygens in the ring. It has the formula \ce{C4H8O2}.
    \end{itemize}
    \item You can have lithium, sodium, potassium, thalium, aluminum, etc. alkyls.
    \item Transition metal alkyls:
    \begin{itemize}
        \item $\beta$-\ce{H} elimination is a problem (this is why TM alkyls were only discovered later).
        \begin{itemize}
            \item An elecronically and coordinatively unsaturated metal ethyl complex reacts in a way that heavily favors the ethylene olefin product.
            \item Note, however, that this reaction is less favored for $d^0$ metals.
        \end{itemize}
        \item First characterized in the early 1900s.
        \begin{itemize}
            \item \ce{I_xPt-Me_{4-x}^2-} and \ce{IPt-Me3^2-} were partially characterized.
        \end{itemize}
        \item First quasi-stable example: \ce{FeCp(CO)2Et} has $18\,\e[-]$. Over time, however, UV light causes it to lose a carbonyl group and form an ethylene olefin hydride (as from a reverse 1,2-migration).
        \item They began to use alkyl ligands with no $\beta$-hydrogens or unreactive $\beta$-hydrogens.
        \begin{itemize}
            \item They made compounds such as \ce{W(CH3)6}, \ce{M-CH2-Ph}, \ce{M-CH2-{$t$-}Bu}, and \ce{M-CH2-TMS}\footnote{Note that TMS is \underline{t}ri\underline{m}ethyl\underline{s}ilyl, a ligand of the structure \ce{SiMe3}.}.
            \item 1-adamantyl, norbornyl have $\beta$-hydrogens, but elimination is disfavored by ring strain.
            \item Fluoroalkyls and metallacycles were also attractive. A fluoroalkyl will not undergo hydride elimination [obviously], but halide elimination is a possible issue. Metallacycles cannot adopt a syn-coplanar arrangement to eliminate.
            \item \ce{M-C#C-H} as well (the $\beta$-hydrogen is once again in the wrong position).
        \end{itemize}
    \end{itemize}
    \item Types of synthesis of transition metal alkyls.
    \begin{enumerate}
        \item Nucleophilic attack on \ce{M}:
        \begin{equation*}
            \ce{$y$R- + MX_y -> MR_y + $y$X-}
        \end{equation*}
        \begin{itemize}
            \item For example, \ce{WCl6 + 6MeLi -> WMe6 + 6LiCl}.
            \item Electron transfer can also be an issue (alkyl lithium agents can be strongly reducing).
            \item There is also the possibility of productive radical mechanisms, but there is lots of side reactivity in radical mechanisms, too. We can limit side reactions with a less reducing nucleophile.
        \end{itemize}
        \item Electrophilic attack on \ce{M}.
        \begin{itemize}
            \item For example, \ce{Mn(CO)5- + MeI -> MnMe(CO)5 + I-}.
            \item Also, \ce{Fp- + Ph2I+ -> CpFe(CO)2Ph + IPh}. Note that \ce{Ph2I+} is an \textbf{arylating reagent}.
        \end{itemize}
        \item Oxidative addition.
        \begin{itemize}
            \item Vaska's complex and \ce{MeI}.
            \item Also \ce{C-H} activation: \ce{M + H-CR3 -> H-M-CR3}.
            \begin{itemize}
                \item This is much more difficult than oxidative addition to alkyl halides and is a big area of interest to organic chemistry.
            \end{itemize}
        \end{itemize}
        \item Insertion of olefins or alkynes.
        \begin{equation*}
            \ce{X-M-|| -> M-CH2-CH2-X}
        \end{equation*}
        \begin{itemize}
            \item X can be an alkyl or a hydride.
            \item \ce{(Et3P)2Pt^{II}ClH ->[||] (Et3P)2PtClEt}.
            \item \ce{(Et3P)2Pt^{II}ClH ->[|||] (Et3P)2PtCl{Vinyl}}.
            \begin{itemize}
                \item Insertion almost always happens with a \emph{trans} disposition to the metal so the \ce{H} that was originally attached to the metal ends up on the vinylic group \emph{cis} to the metal.
            \end{itemize}
            \item Also, \ce{Cp(CO)3MoH ->[N2CH2] Cp(CO)3Mo-CH3} (note that the compound above the arrow is diazomethane, a common \ce{CR2} transfer reagent).
            \item All of these require an open coordination site.
        \end{itemize}
        \item External nucleophilic attack on an olefin.
        \begin{equation*}
            \ce{M-|| +\; \charge{180=\:}{Nu} -> M-CH2-CH2-Nu}
        \end{equation*}
    \end{enumerate}
    \item \textbf{Aryl}: Similar to an alkyl group, but no $\beta$-\ce{H}'s to eliminate.
    \item \ce{M-C} multiple bonds.
    \item Three types of compounds: Carbenes/alkylidenes (\ce{M=CR2}), carbynes/alkylidynes, and carbides.
    \begin{itemize}
        \item We will focus on carbenes today.
    \end{itemize}
    \item Two limiting regimes (for carbenes):
    \begin{figure}[H]
        \centering
        \begin{subfigure}[b]{0.35\linewidth}
            \centering
            \schemestart[-90]
                \chemfig{L_{\emph{n}}M=C(-[1]XR_1)(-[7]R_2)}
                \arrow(.south--.north){<->}
                \chemfig{L_{\emph{n}}\charge{90[yshift=1mm]=$\ominus$}{M}-C(=[1]\charge{90[yshift=1mm]=$\oplus$}{X}R_1)(-[7]R_2)}
            \schemestop
            \caption{Fischer carbenes.}
            \label{fig:carbeneResonancea}
        \end{subfigure}
        \begin{subfigure}[b]{0.35\linewidth}
            \centering
            \schemestart[-90]
                \chemfig{L_{\emph{n}}M=C(-[1]R_1)(-[7]R_2)}
                \arrow(.south--.north){<->}
                \chemfig{L_{\emph{n}}\charge{90[yshift=1mm]=$\oplus$}{M}-\charge{90[yshift=1mm]=$\ominus$}{C}(-[1]R_1)(-[7]R_2)}
            \schemestop
            \caption{Schrock alkylidenes.}
            \label{fig:carbeneResonanceb}
        \end{subfigure}
        \caption{Carbene regime resonance structures.}
        \label{fig:carbeneResonance}
    \end{figure}
    \begin{enumerate}
        \item Fischer carbenes: \ce{L_nM=C(XR1)(R2)}.
        \begin{itemize}
            \item Electrophilic at \ce{C}.
            \item $\ce{X}=\ce{O},\ce{NR},\ce{S}$.
            \item Lower valent metals from the middle-late transition metals.
            \item $\pi$-acceptors, L-type ligands.
        \end{itemize}
        \item Schrock alkylidenes: \ce{L_nM=C(R1)(R2)}.
        \begin{itemize}
            \item Nucleophilic at \ce{C}.
            \item High valent metals from the early-middle transition metals.
            \item $\pi$-donors, \ce{X2}-type ligand.
            \item \ce{R1},\ce{R2} are typically alkyls or aryls.
        \end{itemize}
    \end{enumerate}
    \begin{itemize}
        \item Both are two electron donors.
    \end{itemize}
    \item Synthesis of carbenes:
    \begin{itemize}
        \item Fischer carbenes were made first historically.
        \item For example, \ce{W(CO)6 + MeLi -> Li+[(CO)5W-C(=O)(Me)]- ->[Me3O+BF4-] (CO)5W=C(OMe)(Me)}.
        \item Also, \ce{W(CO)6 + MeLi ->[2KC8][-CO] K2[W(CO)5] ->[Cl-C(=O)-OR][R-X] (CO)5W=C(OMe)(Me)}.
        \begin{itemize}
            \item Note that a carbonyl group is lost after the first step to maintain an electron count of 18, instead of forcing one of 20. Is this method of reducing a compound to remove ligands common?
        \end{itemize}
        \item Also, \ce{W(CO)6 + MeLi ->[2KC8][-CO] K2[W(CO)5] ->[R1-C(=O)-NR2] (CO)5W-C(O-)(R1)(NR2) ->[TMSCl] (CO)5W=C(NR2)(R1)}.
        \item More syntheses listed.
    \end{itemize}
    \item Classic alkylidenes:
    \begin{itemize}
        \item \ce{TaCl5 + $\frac{3}{2}$Np2Zn -> (Np)3TaCl2 ->[2NpLi] Np3Ta=C({$t$-}Bu)(H)}\footnote{Note that we are using \ce{Np} to denote a neopentyl group.}.
        \item \ce{TaCl5 + Np2Zn -> Np2TaCl3 ->[2L] Ta(=C({$t$-}Bu)(H))Cl3L2}.
        \item More syntheses listed.
        \item Essentially, in every synthesis, you make a transiently saturated penta- or hexa-coordinated tantalum center and then do an alkyl elimination to give you the desired alkylidene.
        \item These syntheses were discovered in Dick Schrock's attempts to synthesize \ce{TaMe5}, which actually can't be done because it's so unstable (because of its tendency to participate in bimolecular elimination).
    \end{itemize}
    \item Olefin metathesis (see Figure \ref{fig:olefinMetathesis-W}).
    \begin{figure}[h!]
        \centering
        \schemestart
            \chemfig{W(-Cl)(-[1]Cl)(-[2]Cl)(-[4]MeO)(-[5]MeO)(-[6]OMe)}
            \arrow{->[\small\ce{6ClMgNp}]}[,1.4]
            \chemfig{W(-[2]Np)(~[:-30]-[::0]{\emph{t}-}Bu)(<[:-110]Np)(>:[:-150]Np)}
            \arrow{->[\small\ce{3HCl}][\footnotesize\chemfig[atom sep=1.4em]{MeO-[1]--[7]OMe}]}[,1.8]
            \chemfig{W?(~-{\emph{t}-}Bu)(-[1]Cl)(-[2]Cl)(-[4]O(-[:120])-[:-150,0.6]-[:-60,0.5]-[:-20,0.7]O?(-[:-60]))(-[6]Cl)}
            \arrow{->[][*{0}\footnotesize\hbox{\chemfig[atom sep=1.8em]{**6(---(-(-[2])(-[:-30]))-(-[,,,2]HNTMS)-(-(-[2])(-[:-150]))-)}}]}[-90,2]
            \chemfig{W?(~-{\emph{t}-}Bu)(-[1]Cl)(-[2,,,2]HNAr)(-[4]O(-[:120])-[:-150,0.6]-[:-60,0.5]-[:-20,0.7]O?(-[:-60]))(-[6]Cl)}
            \arrow{->[\small\ce{NEt3}]}[180]
            \chemfig{W?(=_[:30]-[:-30]{\emph{t}-}Bu)(=[2]\chemabove{N}{Ar})(-[4]O(-[:120])-[:-150,0.6]-[:-60,0.5]-[:-20,0.7]O?(-[:-60]))(-[6]Cl)}
            \arrow{->[\small\ce{NOR}]}[180]
            \chemfig{W(=[2]\chemabove{N}{Ar})(=[:-30]X)(<[:-110]RO)(>:[:-150]RO)}
        \schemestop
        \caption{Tungsten olefin metathesis.}
        \label{fig:olefinMetathesis-W}
    \end{figure}
    \begin{itemize}
        \item We'll talk about what this is later.
        \item Done with tungsten and molybdenum in Dick Schrock's lab as well.
        \item The second step contains an alkylidyne that is very resilient throughout the rest of the mechanism.
    \end{itemize}
    \item Grubbs type alkylidenes are synthetically much more useful and ruthenium based.
    \item Tebbe's reagents:
    \begin{figure}[H]
        \centering
        \schemestart
            \chemfig{Cp_2TiCl_2}
            \+
            \chemfig{Me_3Al}
            \arrow
            \chemfig{Cp_2Ti?=CH_2-[6,,1]Al(-Me)(-[6]Me)-[4]Cl?[,,dashed]}
        \schemestop
        \caption{Tebbe's reagents.}
        \label{fig:tebbeReagent}
    \end{figure}
    \begin{itemize}
        \item If you treat a Tebbe's reagent with an olefin in the presence of DMAP, you can form a 4-membered ring.
    \end{itemize}
    \item N-heterocyclic carbenes (NHC complexes).
    \begin{itemize}
        \item Neutral L-type ligands that are very strong $\sigma$-donors (much stronger than phosphines), but very weak $\pi$-acceptors.
    \end{itemize}
    \item Reactivity:
    \begin{itemize}
        \item $2+2$ addition: The defining reactivity of carbenes.
        \begin{itemize}
            \item React a carbene with an olefin to make a metallacycle, which can then collapse into olefin metathesis.
            \item It allows us to swap the two \ce{CR2} fragments, which is pretty useful.
        \end{itemize}
        \item Cyclopropanation: Occurs through electrophilic, Fischer carbenes.
        \begin{itemize}
            \item Allows us to create a cyclopropane compound with both of the carbene R groups attatched to one of the cyclopropane carbons.
        \end{itemize}
    \end{itemize}
    \item \textbf{Carbide}: A terminal carbon atom; a \ce{C^4-} ligand.
\end{itemize}



\section{Lecture 9: Olefins, Carbonyls, and Phosphines}
\begin{itemize}
    \item \marginnote{4/16:}Top $\pi$-acceptors: \ce{CO}, \ce{NO+}, and \ce{PF3}.
    \begin{itemize}
        \item \ce{CO} is also a decent $\sigma$ donor.
    \end{itemize}
    \item \ce{CO} provides a great IR handle, enabling characterization.
    \begin{itemize}
        \item $\nu_{\ce{CO}}$ is a measure of the electron density on the metal center.
    \end{itemize}
    \item We generally consider \ce{CO} in the resonance structure\ \ \chemfig{\charge{180=\:,90[yshift=1mm]=$\ominus$}{C}~\charge{0=\:,90[yshift=1mm]=$\oplus$}{O}}\ .
    \begin{itemize}
        \item Each atom in \ce{CO} has four $\pi^*$ lobes in the plane perpendicular to the bond axis and slightly oriented away from those on the other atom (note that the lobes on the carbon are significantly larger than those on the oxygen). Additionally, there is a nonbonding $\sigma$ orbital originating from atomic $p$ orbitals that runs along the bond axis.
        \item Recall the \ce{CO} resonance structures.
    \end{itemize}
    \item Free \ce{CO} has a stretching frequency of $\nu_{\ce{CO}}=\SI{2143}{\per\centi\meter}$.
    \begin{itemize}
        \item This is also a range of the spectrum with little else going on, making these peaks easily identifiable.
        \item \ce{CO} has a large dipole (about $\mu=\SI{4.80}{D}$). This also really helps it stick to transition metals.
    \end{itemize}
    \item In \ce{Mn(CO)5Me}, the \ce{Mn-Me} bond is $\SI{2.18}{\angstrom}$ and the \ce{Mn-CO} bonds are $\SI{1.86}{\angstrom}$, so these latter bonds clearly have some multiple bond character.
    \begin{itemize}
        \item The trans effect is also partially at play.
    \end{itemize}
    \item Synthesis:
    \begin{itemize}
        \item In 1890, the Mond process is discovered: \ce{NH3 + NaCl + CO2 -> Na2CO3 + NH4Cl}.
        \begin{itemize}
            \item After a year however, the plant built to run this process was trashed! The nickel in the steel pipes was reacting with \ce{CO} to form \ce{Ni(CO)4}, a volitle compound.
            \item Leaching the nickel weakened the pipes, causing them to fall apart.
            \item Take away: metal-carbonyl complexes can be pretty easy to form (as a general synthetic scheme, just add \ce{CO} and apply pressure, as in the following mechanisms).
        \end{itemize}
        \item Pressurized \ce{CO}:
        \begin{itemize}
            \item \ce{Co + CO ->[$\SI{35}{atm}$][RT] Co2(CO)8}\footnote{RT is short for room temperature.}.
            \item \ce{Fe + CO ->[$\SI{300}{atm}$][$\SI{300}{\celsius}$] Fe(CO)5}.
        \end{itemize}
        \item Reduce with \ce{CO}:
        \begin{itemize}
            \item \ce{Re2O7 ->[CO] Re2(CO)10 + CO2}. \ce{CO2} is given off by reductive elimination from an intermediate.
            \item More syntheses listed.
        \end{itemize}
    \end{itemize}
    \item We predict stoichiometry (i.e., constitutional isomerism) with the $18\,\e[-]$ rule (except for square planar).
    \begin{itemize}
        \item Group 9: \ce{M4(CO)12} is common.
        \item Group 8: \ce{M3(CO)12} is common.
    \end{itemize}
    \item IR signatures.
    \begin{itemize}
        \item Ketones are usually $\num{1750}$-$\SI{1720}{\per\centi\meter}$.
        \item \ce{R3COH} compounds are usually $\SI{1100}{\per\centi\meter}$.
        \item $\nu_{\ce{CO}}$ will increase if there is no backbonding, and will decrease if there is strong backbonding.
    \end{itemize}
    \item Influences on backbonding:
    \begin{itemize}
        \item Primarily the metal oxidation state and electronegativity.
        \item Electron count does not matter as much, aside from $d^0$ (obviously no backbonding can occur in this case).
        \item Second and third row metals tend to backbond more strongly (stems from their larger orbital radius).
    \end{itemize}
    \item Thus, we can predict that low-valent, early metals are strong backbonders, and high-valent, late metals are weak backbonders.
    \item "Extreme" carbonyls (those with a high degree of backbonding, resulting from a very reduced/comparably early metal center):
    \begin{itemize}
        \item \ce{Fe(CO)5 ->[2Na / THF][-CO] Na2Fe(CO)4^2-}.
        \item \ce{M(CO)_n ->[2Na][-CO] Na2M(CO)_{n-1}^2-}.
        \item \ce{Re2(CO)10 ->[Na] Na+Re(CO)5-}.
        \item \ce{Re2(CO)10 ->[6Na / HMPA][-2CO] Na3Re(CO)4^3-}.
        \item \ce{Mo(CO)6 ->[2Na] Mo(CO)5^2-}.
        \item \ce{Mo(CO)6 ->[$\Delta$ TMEDA] (CO)4Mo(EDA) ->[4NaNH3] Na4Mo(CO)4^4-}.
        \begin{itemize}
            \item This is $d^{10}$ and thus $\nu_{\ce{CO}}<<<\SI{2000}{\per\centi\meter}$.
            \item TMEDA is tetramethylethylenediammine.
        \end{itemize}
    \end{itemize}
    \item We can also go the other direction to find carbonyls with a very low degree of backbonding.
    \begin{figure}[h!]
        \centering
        \schemestart
            \chemleft{[}
                \chemfig{Zr(-[3]Cp)(-[5]Cp)(-[7]CO)(-[:30](=[2]O)-[:-30])}
            \chemright{]^+}
            \arrow{<=>}
            \chemleft{[}
                \chemfig{Zr(-[3]Cp)(-[5]Cp)(-[7]CO)(-[:30](-[2])=[:-30]O)}
            \chemright{]^+}
        \schemestop
        \caption{Low backbonding carbonyl complexes.}
        \label{fig:extremeCarbonyl}
    \end{figure}
    \begin{itemize}
        \item The reactant has $\nu_{\ce{CO}}=\SI{2123}{\per\centi\meter}$, which is slightly lower than that of free \ce{CO} due to some $\sigma$ donation from the acetyl ligand.
        \begin{itemize}
            \item The product has $\nu_{\ce{CO}}=\SI{2176}{\per\centi\meter}$, which is slightly higher because the acetyl's oxygen is pushing more electron density off of the \ce{CO} ligand.
        \end{itemize}
        \item Higher $d$ counts can also lead to higher stretching frequencies.
        \begin{itemize}
            \item It may seem counterintuitive that compounds with more electrons backbond less, but backbonding is actually dictated by reduction potentials, electronegativity, and oxidation states.
            \item Indeed, both \emph{cis}- and \emph{trans}-\ce{PtCl2(CO)2} exhibit \ce{CO} stretching frequencies that are very similar to that of free \ce{CO}.
        \end{itemize}
        \item More extreme: \ce{2IrF6 + 15CO + HSbF5 ->[SbF5][$\SI{60}{\celsius}$, $\SI{12}{\hour}$] 2[Ir(CO)6]^3+}.
        \begin{itemize}
            \item This gives us $\nu_{\ce{CO}}=\SI{2254}{\per\centi\meter}$.
            \item We also have \ce{Au(CO)2+} at $\SI{2217}{\per\centi\meter}$.
        \end{itemize}
        \item Reducing extreme:
        \begin{itemize}
            \item \ce{Zr(CO)6^2-} is $\SI{1757}{\per\centi\meter}$.
            \item \ce{W(CO)6} is $\SI{1983}{\per\centi\meter}$.
        \end{itemize}
    \end{itemize}
    \item Neutral L-type olefins: \ce{M-||}.
    \item Dewar-Chatt-Duncanson model: Very related to Figure \ref{fig:oxAdn-concerted-orbitals}.
    \begin{itemize}
        \item Thus, the ethylene is both a $\sigma$ donor and $\pi$ acceptor.
        \begin{itemize}
            \item This relates it to \ce{CO} as a ligand. Of course, \ce{CO} is $C_{\infty v}$ and \ce{CO} has 2 $\pi$-accepting orbitals instead of just one, but there is a relation nonetheless.
        \end{itemize}
    \end{itemize}
    \item Use ionization potential to measure the magnitude of $\sigma$ donation:
    \begin{itemize}
        \item Ethylene: $\SI{10.5}{eV}$.
        \item \ce{CO}: $\SI{14}{eV}$.
    \end{itemize}
    \item Thus, ethylene is a better $\sigma$ donor but worse $\pi$ acceptor.
    \item Comparing \ce{C=C} bond length in Zeise's salt (see Figure \ref{fig:ZeisesSalt}), \ce{(Ph3P)2Pt-||}, and \ce{C2H4}.
    \begin{itemize}
        \item Bond length is significantly elongated (close to a \ce{C-C} single bond) for \ce{(Ph3P)2Pt-||} but similar for the other two.
        \item We can also see that the \ce{C-H} bonds are bent back by $\ang{15}$ in \ce{(Ph3P)2Pt-||}, which is consistent with $sp^3$ hybridization.
    \end{itemize}
    \item There are two resonance structures for an olefin: An olefin adduct and a metallacyclopropane.
    \begin{itemize}
        \item The olefin adduct is an L-type ligand and the metallacyclopropane is an \ce{X2}-type ligand, but by convention we almost always treat olefins as L-type ligands.
        \item Both are $2\,\e[-]$ donors.
    \end{itemize}
    \item Polyolefins.
    \item Example: Butadiene.
    \begin{figure}[h!]
        \centering
        \chemfig{M-[2]-[:-157,0.85,,,white]=_[1]-=^[5]}
        \caption{Butadiene.}
        \label{fig:butadiene}
    \end{figure}
    \begin{itemize}
        \item We have four bonding modes: $\sigma$, $\pi$, $\pi^*$, and $\delta^*$.
        \begin{itemize}
            \item Each corresponds to a different orientation of the $p_z$ orbitals at each carbon and has a different number of nodes.
        \end{itemize}
        \item $\pi$ and $\pi^*$ orbitals are the major contributors.
        \begin{itemize}
            \item The metal $d$ orbitals will project toward butadiene in a $\pi$-symmetry fashion, and the $s$ orbitals (which would bond with the butadiene $\sigma$ orbital) are more secondary for a transition metal.
            \item $\sigma$ contributions are reduced because of symmetry.
        \end{itemize}
        \item $\sigma$ and $\pi$ orbitals are $\sigma$ donors; $\pi^*$ engages in backbonding.
        \item Lowering the energy of the $\pi$ orbital and populating the $\pi^*$ orbital favors new resonance structures that distort the bonds in butadiene (why to the resonance structure shown?).
        \begin{itemize}
            \item Originally, the bonds are distorted. However, after bonding, all bonds in butadiene are the same length.
        \end{itemize}
    \end{itemize}
    \item Review of insertion, nucleophilic attacks, and electrophilic attacks.
    \item \ce{M#C-OE} where \ce{E} is an electrophile is a Fischer carbyne.
    \item The relative backbonding plays a crucial role in determining reactivity.
    \item Olefin limiting cases: Metallacyclopropane and direct $\pi$-bonding.
    \begin{itemize}
        \item Different resonance structures are more suitable for different kinds of attacks.
        \item The stronger the backbonding, the easier it is to get the ligands to play nice with electrophiles.
    \end{itemize}
    \item Orbitals in bridging \ce{CO}.
    \begin{itemize}
        \item A bridging \ce{CO} is a $\sigma$ donor, and donates into $d_{xy}$ metal orbitals.
        \item It's also a $\pi$ acceptor, and accepts electrons from $d_{x^2-y^2}$ orbitals.
        \item Bridging is typically much more activated (in terms of \ce{CO} stretching frequencies).
    \end{itemize}
    \item \ce{CO} activation.
    \begin{itemize}
        \item All renewables technology boils down to finding ways to reduce \ce{CO} into multiple carbon products.
        \item \ce{C-O} $\text{BDE}=\SI[per-mode=symbol]{1075}{\kilo\joule\per\mole}$\footnote{Note that BDE is short for bond dissociation energy.}.
        \begin{itemize}
            \item This is an even higher BDE than that of the \ce{N#N} triple bond.
            \item \ce{CO} is easier to activate due to the dipole (more on this on Canvas).
        \end{itemize}
        \item There is a lot of interest in deoxygenating \ce{CO} and forming \ce{C-C} bonds from the products. The classic way to do this used highly reduced transition metal centers:
        \begin{equation*}
            \ce{Ta^{III}(OSi({$t$-}Bu)3)3 ->[2CO] 2R3SiO3Ta^V=O + (R3SiO)3Ta=C=C=Ta(OSiR3)3}
        \end{equation*}
        \begin{itemize}
            \item The latter product exhibits significant variation in its double bond lengths (the \ce{C=C} one is much shorter than the two \ce{Ta=C} ones).
        \end{itemize}
    \end{itemize}
    \item Lists a few more syntheses.
    \item Don't worry about memorizing all these reactions; just know the general types and reagents.
\end{itemize}



\section{Midterm 1 Review}
\begin{itemize}
    \item \marginnote{4/17:}Up through what lecture is covered on the midterm?
    \begin{itemize}
        \item Up through lecture 11.
    \end{itemize}
    \item For problem 7: Do you want us to count an olefin as having coordination number 1 or 2? Is it ok to treat the olefin as if it's ethylene.
    \begin{itemize}
        \item An olefin has coordintion number 2.
    \end{itemize}
    \item Are alkyl groups more stable?
    \item Synthesis questions.
    \begin{itemize}
        \item Try to stay under $18\,\e[-]$.
        \begin{itemize}
            \item Don't include the electron counts as part of the synthesis but think about them for yourself.
            \item \ce{ZrCl4} has $8\,\e[-]$? We describe metal chlorides as simply as possible, but it's probably in a chain or something in real life so each \ce{Zr} gets more electrons.
        \end{itemize}
        \item \ce{CO} just comes off (it's a gas).
        \item You can just say "add a CpMe" group and some \ce{CO}s will come off.
        \item We can do $-\ce{CO}$ as needed.
        \item Keep in mind which metal centers are more labile based on their oxidation state.
        \item We don't have to draw structures; we can just use formulas.
        \item Difference in oxidation state of the zirconiums means we need to add a reductant at some point. We can choose to reduce first or add Cp's first. We'll add Cp's first because adding the Cp's might make it more stable before we reduce it.
        \item It's generally lithium or sodium, not potassium.
        \item As reductants, we can use \ce{KC8} (potassium graphite), \ce{Na^0} (sodium metal), or \ce{NaNapth} (sodium napthalanide), but not hydride reductants like in orgo.
        \begin{itemize}
            \item In the lab, you don't want to use too strong a reductant.
            \item However, in this theoretical case, it's ok to use these real hammers.
            \item Some mild reductants: \ce{CoCp2} (cobaltacene) and \ce{CoCp^*}.
        \end{itemize}
        \item We reduce before adding \ce{CO}s because if we added two \ce{CO}s right now, we'd get a $20\,\e[-]$ species.
        \item \emph{In situ} means in the situation. Anything you do in the same mixture is something you've done \emph{in situ}.
        \item Oxidants:
        \begin{itemize}
            \item \ce{AgPF6}: \ce{Ag^0} is super stable?
            \item Bromine can be an oxidant? Look back at orgo notes for dinuclear oxidative addition.
        \end{itemize}
    \end{itemize}
    \item Review the powerpoint on carbonyls.
    \item Midterm tips:
    \begin{itemize}
        \item \ce{CO} is a really important ligand to understand: Know it's MO diagram, resonance structures, IR stretch influences.
        \item Know phosphines: Tolman parameter and cone angle, donor properties, orbitals, etc.
        \item Know some about olefins, too.
        \item Synthesis is very important. Use your electron counting skills, understanding of lability, and redox reactions.
        \item Be able to draw the carbenes and various binding modes of the different ligands we've discussed, including resonance.
        \item Know lability abilities and how to predict orbital splitting and spin state.
        \item Know the two equations from magnetism.
        \item Different types of insertion and their final stereochemistry.
        \item If there's a drawing in the notes, ask yourself if you know how to draw it.
    \end{itemize}
    \item Homework 2 should be a pretty good review for Midterm 1.
    \item Question 11:
    \begin{itemize}
        \item For favorability, talk about hard/soft.
    \end{itemize}
    \item Question 13:
    \begin{itemize}
        \item Part a: Draw the starting compound and the products of the \ce{MeI} and \ce{H2} reactions.
        \item Part c: The compounds with each type of \ce{X} (i.e., all of the halogens).
    \end{itemize}
\end{itemize}



\section{Lecture 10: Nitrosyls, Allyls, and More}
\begin{itemize}
    \item \marginnote{4/19:}First off, a survey of other $\pi$ acceptors.
    \begin{itemize}
        \item Many of these are analogous to \ce{CO}.
    \end{itemize}
    \item \ce{C#S} (carbon sulfide).
    \begin{itemize}
        \item $2\,\e[-]$ L-type donor.
        \item Not stable unless it's ligated to a metal (will decompose as a free gas).
        \item Due to the weak \ce{C-S} binding, it can be more of a $\pi$ acceptor than \ce{CO} even though the polarization is a bit off.
    \end{itemize}
    \item \ce{[C#N]-} (cyanide).
    \begin{itemize}
        \item $1\,\e[-]$ X-type.
        \item Much better $\sigma$ donor, but a much weaker $\pi$ acceptor.
    \end{itemize}
    \item \ce{C#N-R} (isocyanides).
    \begin{itemize}
        \item Quite similar to \ce{CO} (via the isolobal/isoelectronic analogy).
        \item Slightly stronger $\sigma$ donor, but a slightly weaker $\pi$ acid.
    \end{itemize}
    \item \ce{PR3} (phosphines).
    \begin{itemize}
        \item Good $\sigma$ donors (better than \ce{CO}), and ok $\pi$ acceptors.
        \item $\pi$ acceptance occurs through the $\sigma^*$ orbitals.
        \begin{itemize}
            \item More polarized the \ce{P-R} bond $\Rightarrow$ more $\pi^*$ character on the phosphorous $\Rightarrow$ better $\pi$ acceptor \ce{PR3} will be.
            \item Phosphines with strong electron-donating R groups will be better $\sigma$ donors but worse $\pi$ acceptors.
        \end{itemize}
        \item Know \ce{P(OR)3} ligands (phosphites).
        \begin{itemize}
            \item Very good $\pi$ acceptors.
        \end{itemize}
        \item \ce{PF3} is almost as good as \ce{CO} in terms of its $\pi$-acceptance.
    \end{itemize}
    \item \ce{[N#O]+} (nitrosyls).
    \begin{itemize}
        \item Isoelectronic/isolobal to \ce{CO}, so it engages in the same types of interactions.
        \item Probably the strongest $\pi$ acceptor known.
        \item Two main modes of binding (linear and bent).
        \item Bioinorganic chemistry of \ce{NO+}: An important signaling molecule; a potent vasodilator (i.e., a molecule that when released in the bloodstream causes blood vessels to expand); important in blood pressure.
    \end{itemize}
\end{itemize}
\begin{tchart}{1.4}{Linear}{Bent}
    \setcharge{extra sep=2.5pt}\chemfig[atom sep=1.2em]{M-\charge{90[yshift=1mm]=$\oplus$}{N}~\charge{90[yshift=1mm]=$\oplus$,0=\:}{O}} & \setcharge{extra sep=2.5pt}\chemfig[atom sep=1.2em]{M-[:30]\charge{90=\:}{N}=[:-30]\charge{0=\:,-90=\:}{O}}\rule{0pt}{0.9cm}\\
    LX-type. & X-type.\\
    $3\,\e[-]$ donor. & $1\,\e[-]$ donor.\\
    3 CO's provide similar stability to 2 NO's (in terms of electron counting). & \\
    Oxidation state: Formally \ce{NO+} but this is not a good description (it would imply \ce{M^-}). & Oxidation state: Formally anionic, so \ce{M^+}.\\
\end{tchart}
\begin{itemize}
    \item Oxidation state debates led to the development of \textbf{Enemark-Feltham notation}.
    \item \textbf{Enemark-Feltham notation}: Considers the $d$ count of the metal center and any extra electrons beyond \ce{NO+} that the \ce{NO} brings.
    \item Consider heme centers.
    \begin{itemize}
        \item Hemes bind to porphyrins.
        \begin{itemize}
            \item Porphyrins are the macrocyclic aromatic ligands that bind iron and magnesium in a fashion hemoglobin and chlorophyll, respectively.
            \item Porphyrins are very common in biology in general, too.
        \end{itemize}
        \item Formally a square-planar binding ligand with a $2^-$ charge when deprotonated.
        \item Thus, heme iron is $2+$, i.e., $d^6$.
        \item If heme \ce{FeNO} is neutral, \ce{*NO} bound as a radical: \ce{\{FeNO\}^7}.
        \item If heme \ce{FeNO-} is anionic, \ce{NO-} bound as an anion: \ce{\{FeNO\}^8}.
        \item If heme \ce{FeNO+} is cationic, \ce{NO+} bound as a cation: \ce{\{FeNO\}^6}.
    \end{itemize}
    \item Nitrosyl geometry (linear vs. bent) is sensitive to the $d$ count of the metal.
    \begin{itemize}
        \item If the metal wants to get an extra $3\,\e[-]$, \ce{NO+} will be linear. Otherwise, \ce{NO+} will be bent. 
        \item An example of a limiting case: When bonding to benzene, \ce{NO+} will be bent because there is no possible $\pi$ bonding.
    \end{itemize}
    \item Synthesis of nitrosyls:
    \begin{itemize}
        \item Two main routes:
        \begin{gather*}
            \ce{M^0 + NO+X- -> [M-NO]+X-}\\
            \ce{M^0 + *NO -> M-NO}
        \end{gather*}
        \begin{itemize}
            \item In the first case, we react a reduced metal center with a nitrosonium salt to in-situ oxidize the metal center to make a metal-nitrosyl cation with the \ce{X-} as a counteranion.
            \item In the second case, we react a metal center with \ce{NO*} gas, which is a radical and pretty nasty.
        \end{itemize}
        \item One less common route:
        \begin{gather*}
            \ce{R-O-N=O ->[H+] ROH + NO+}
        \end{gather*}
        \begin{itemize}
            \item An analogous reaction with metals is \ce{M-N(=O)-O ->[PR3] M-NO + R3PO}.
        \end{itemize}
    \end{itemize}
    \item Note that in general, because \ce{NO} and \ce{CO} are both such strong $\pi$ acceptors, they do not go \emph{trans} to each other.
    \begin{itemize}
        \item One notable exception: \ce{V(CO)5(NO)}.
    \end{itemize}
    \item Allyls:
    \begin{figure}[H]
        \centering
        \begin{subfigure}[b]{0.3\linewidth}
            \centering
            \chemfig{=_[:30]-[:-30]\charge{-90[yshift=-1.5mm]=$\ominus$}{}}
            \vspace{1em}
            \caption{The ligand.}
            \label{fig:ligand-Allyla}
        \end{subfigure}
        \begin{subfigure}[b]{0.3\linewidth}
            \centering
            \begin{tikzpicture}
                \node [yshift=3cm] {\chemfig{(-[2,0.15]{-})-[:30](-[2,0.15]+)-[:-30](-[2,0.15]{-})}};
                \node [yshift=1.5cm] {\chemfig{(-[2,0.15]+)-[:30]-[:-30](-[2,0.15]{-})}};
                \node [yshift=0cm] {\chemfig{(-[2,0.15]+)-[:30](-[2,0.15]+)-[:-30](-[2,0.15]+)}};
    
                \draw [dashed] (-1,2.25) -- ++(2,0);
                \draw [dashed] (-1,0.75) -- ++(2,0);
    
                \footnotesize
                \node [anchor=west] at (2,3) {Totally antibonding};
                \node [anchor=west] at (2,1.5) {Totally nonbonding};
                \node [anchor=west] at (2,0) {Totally bonding};
            \end{tikzpicture}
            \caption{The $\pi$ system.}
            \label{fig:ligand-Allylb}
        \end{subfigure}
        \caption{The allyl ligand.}
        \label{fig:ligand-Allyl}
    \end{figure}
    \begin{itemize}
        \item $3\,\e[-]$ LX-type donor.
        \item Barry Trost's palladium allyls: Changing side-on vinyl ethers into allyls (structures shown).
    \end{itemize}
    \item Synthesis of allyls:
    \begin{itemize}
        \item We can synthesize allyls (from olefins) that bind to their metal center from above and below the molecular plane.
        \item We can synthesize allyls by oxidizing a butadiene with acid.
        \item There also exist allyls in the form of isobutane with conjugated electron systems between each carbon. This bonds to metal centers from above.
        \item More syntheses listed.
    \end{itemize}
    \item Note: Sodium is a possible reductant (\ce{Mn2(CO)10 ->[Na] 2Mn(CO)5-}).
    \item Allyl reactivity:
    \begin{figure}[h!]
        \centering
        \begin{subfigure}[b]{0.9\linewidth}
            \centering
            \begin{tikzpicture}
                \node{
                    \schemestart
                        \chemfig{M-\phantom{i}-[:-100,0.7,,,white]-[:60]-[:120]}
                        \arrow{->[\small\ce{Nu-}]}
                        \chemleft{[}
                            \chemfig{M-\phantom{i}-[2,0.35,,,white]=_[6]-[:-60]-[6]Nu}
                        \chemright{]^-}
                    \schemestop
                };
                \draw [dashed] (-2,-0.56) arc (-35:35:1);
            \end{tikzpicture}
            \caption{Nucleophilic attack.}
            \label{fig:allylReactivitya}
        \end{subfigure}
        \vspace{1em}
        \begin{subfigure}[b]{0.9\linewidth}
            \centering
            \begin{tikzpicture}
                \node{
                    \schemestart
                        \chemfig{M-\phantom{i}-[:-100,0.7,,,white]-[:60]-[:120]}
                        \arrow{->[\small\ce{E+}]}
                        \chemleft{[}
                            \chemfig{M-\phantom{i}-[2,0.35,,,white]=_[6]-[:-60]-[6]E}
                        \chemright{]^+}
                    \schemestop
                };
                \draw [dashed] (-1.95,-0.56) arc (-35:35:1);
            \end{tikzpicture}
            \caption{Electrophilic attack.}
            \label{fig:allylReactivityb}
        \end{subfigure}
        \vspace{1em}
        \begin{subfigure}[b]{0.49\linewidth}
            \centering
            \begin{tikzpicture}
                \node{
                    \schemestart
                        \chemfig{M-\phantom{i}-[:-100,0.7,,,white]-[:60]-[:120]}
                        \arrow{->[\small\ce{CO2}]}
                        \chemfig{M-O-[:30](=[2]O)-[:-30]-[:30]=_[:-30]}
                    \schemestop
                };
                \draw [dashed] (-2.53,-0.56) arc (-35:35:1);
            \end{tikzpicture}
            \caption{Insertion (1).}
            \label{fig:allylReactivityc}
        \end{subfigure}
        \begin{subfigure}[b]{0.49\linewidth}
            \centering
            \begin{tikzpicture}
                \node{
                    \schemestart
                        \chemfig{M-\phantom{i}-[:-100,0.7,,,white]-[:60]-[:120]}
                        \arrow{->[\small\ce{CO}]}
                        \chemfig{M-[:30](=[2]O)-[:-30]-[:30]=_[:-30]}
                    \schemestop
                };
                \draw [dashed] (-2.05,-0.56) arc (-35:35:1);
            \end{tikzpicture}
            \caption{Insertion (2).}
            \label{fig:allylReactivityd}
        \end{subfigure}
        \vspace{1em}
        \begin{subfigure}[b]{0.9\linewidth}
            \centering
            \schemestart
                \chemfig{M(-[1]\phantom{i}-[1,0.3,,,white]-[::-120,,,,white]@{1C1}-[::180]-[::60]@{1C2}-[::-60]R)(-[7]\phantom{i}-[7,0.3,,,white]-[::120,,,,white]@{2C1}-[::180]-[::-60]@{2C2}-[::60]R)}
                \arrow{->}
                \chemfig{M^0}
                \arrow{0}[,0]\+{,,1.6cm}
                \chemfig{R-[2]=_[:150]-[2]=_[:30]-[2]R}
            \schemestop
            \chemmove{
                \draw [dashed,-] ([xshift=-2pt,yshift=-0.6pt]1C1.center) to[bend right=30] ([xshift=-2pt,yshift=-2pt]1C2.center);
                \draw [dashed,-] ([xshift=-2pt,yshift=0.6pt]2C1.center) to[bend left=30] ([xshift=-2pt,yshift=2pt]2C2.center);
            }
            \caption{Coupling/reductive elimination.}
            \label{fig:allylReactivitye}
        \end{subfigure}
        \caption{Allyl reactivity.}
        \label{fig:allylReactivity}
    \end{figure}
    \item Other extended $\pi$ systems.
    \item Cyclopentadienyl (Cp).
    \begin{figure}[h!]
        \centering
        \begin{subfigure}[b]{0.3\linewidth}
            \centering
            \chemfig{[:18]*5(-=-\charge{90[yshift=1.5mm]=$\ominus$}{}-=)}
            \caption{The ligand.}
            \label{fig:ligand-Cpa}
        \end{subfigure}
        \begin{subfigure}[b]{0.6\linewidth}
            \centering
            \begin{tikzpicture}
                \node [yshift=4cm] at (-1.5,0) {\chemfig{[:18]*5((-[,0.15]{-})-(-[,0.15]{-})-(-[,0.15]+)-(-[,0.15]{\color{white}-})-(-[,0.15]+)-)}};
                \node [yshift=4cm] at (1.5,0)  {\chemfig{[:18]*5((-[,0.15]{-})-(-[,0.15]{-})-(-[,0.15]+)-(-[,0.15]{-})-(-[,0.15]+)-)}};
                \node [yshift=2cm] at (-1.5,0) {\chemfig{[:18]*5((-[,0.15]+)-(-[,0.15]{-})-(-[,0.15]{-})-(-[,0.15]{\color{white}+})-(-[,0.15]+)-)}};
                \node [yshift=2cm] at (1.5,0)  {\chemfig{[:18]*5((-[,0.15]{-})-(-[,0.15]{-})-(-[,0.15]+)-(-[,0.15]+)-(-[,0.15]+)-)}};
                \node [yshift=0cm] at (0,0)    {\chemfig{[:18]*5((-[,0.15]+)-(-[,0.15]+)-(-[,0.15]+)-(-[,0.15]+)-(-[,0.15]+)-)}};
    
                \draw [dashed] (-2.5,3) -- ++(5,0);
                \draw [dashed] (-2.5,1) -- ++(5,0);
    
                \footnotesize
                \node [anchor=west] at (3,4) {$\pi$ acceptor};
                \node [anchor=west] at (3,2) {$\pi$ donor};
                \node [anchor=west] at (3,0) {$\sigma$ donor};
            \end{tikzpicture}
            \caption{The $\pi$ system.}
            \label{fig:ligand-Cpb}
        \end{subfigure}
        \caption{The cyclopentadienyl ligand.}
        \label{fig:ligand-Cp}
    \end{figure}
    \begin{itemize}
        \item Strong field ligand.
        \item $5\,\e[-]$ \ce{L2X}-type donor.
    \end{itemize}
    \item Ferrocene is a classic example of a Cp-containing compound.
    \item In 1951, Pauson (very respected, famous chemist) showed that a \ce{Cp-MgBr} Grignard (a Cp anion source) becomes a radical when treated with an oxidant and then dimerizes. This product when heated would lose \ce{H2} and become a fulvalene.
    \begin{itemize}
        \item However, if you use \ce{FeCl3} as the oxidant, you get a compound with the formula \ce{(C5H5)2Fe} (as determined by elemental analysis, or EA).
        \item This compound had a number of interesting properties, all demonstrating that it was very stable.
        \begin{itemize}
            \item Orange solid.
            \item Air stable.
            \item Temperature stable.
            \item Base stable (at $10\%\,\ce{NaOH}$).
            \item Melting point of $\SI{173}{\celsius}$ (thus, a clean melting point transition).
            \item Stable up to $\SI{500}{\celsius}$.
            \item Water stable.
        \end{itemize}
        \item To characterize it as a sandwich compound, scientists used the fact that it has 1 $\nu_{\ce{C-H}}$ stretch in the IR spectrum, 1 peak in \ce{{}^1H} (proton) NMR, and no Diels-Alder reactivity.
        \item Pauson missed this in his initial analysis. Woodward and E. O. Fischer proposed the real structure, which was later confirmed by X-ray diffraction (XRD) crystallography.
    \end{itemize}
    \item There are tons of Cp complexes.
    \item However, almost any $\pi$ system can bond to a metal center in a face-on fashion. For example, there are\dots
    \begin{itemize}
        \item Benzene and tropylium (cycloheptatrienyl cation) aromatic adducts.
        \item Butadiene and borazine nonaromatic adducts.
    \end{itemize}
    \item \textbf{Ring-slipping}: When an aromatically bonded ring redistributes its electron density to localize it and only bind through some of its ring.
    \begin{itemize}
        \item For example, \ce{Ru($\eta^6${-}C6H6)2 -> Ru($\eta^6${-}C6H6)($\eta^4${-}C6H6)}. The driving force is a $20\,\e[-]$ to $18\,\e[-]$ transition.
    \end{itemize}
    \item Triple bonds:
    \begin{itemize}
        \item $\sigma$ donation, (weak) $\pi$ donation, and $\pi$ acceptance (same idea, orbitally, as answer to Homework 2, problem 11a).
        \item Upon binding, we see lengthening of the \ce{C#C} bond and the \ce{C-H} bonds bending backwards away from the metal center.
    \end{itemize}
    \item Phosphines (and derivatives):
    \begin{itemize}
        \item \ce{PR3}: Bulky, soft, good donors, strong field ligands.
        \item \ce{P(OR)3}: Harder (due to the oxygens), weaker donors, stronger acceptors.
        \item Compares basicity in a number of phosphines (phosphines with a greater ability to delocalize the electron pair on the phosphorous are weaker bases [i.e., \ce{PPh3} is weak while \ce{PMe3} is strong]).
    \end{itemize}
    \item \textbf{Tolman electronic parameter}: The $\nu_{\ce{CO}}$ value for \ce{(CO)3NiL} where \ce{L} is some ligand.
    \begin{itemize}
        \item Allows us to guage the $\pi$-accepting character of different ligands: Stronger $\pi$ acceptors will invoke a higher $\nu_{\ce{CO}}$ value.
        \item Weakly basic ligands do not do much $\sigma$ or $\pi$ donation and thus do a lot of $\pi$ acceptance. It follows that these ligands will have a higher Tolman electronic parameter.
        \item There is a good Wikipedia page for this.
        \item Judges $\sigma$/$\pi$ donation/acceptance properties all together (i.e., their ratio), not each individually.
        \begin{itemize}
            \item For example, a ligand that is a strong $\sigma$ donor and a strong $\pi$ acceptor will have a middling Tolman electronic paramenter, whereas a ligand that is a weak $\sigma$ donor and a strong $\pi$ acceptor will have a high Tolman electronic parameter.
        \end{itemize}
    \end{itemize}
    \item $\pi$-accepting character of various ligands:
    \begin{equation*}
        \ce{NMe2}
        < \ce{NH3}
        << \ce{py}
        < \ce{PR3}
        < \ce{P(OR)3}
        << \ce{PF3}
        < \ce{PCl3}
        <<< \ce{CO}
    \end{equation*}
    \begin{itemize}
        \item Ranked on a scale of 0-100 where the $\pi$-accepting character of \ce{CO} is defined to equal 100.
    \end{itemize}
    \item \textbf{Tolman cone angle}: The angle for the cone that encapsulates all of the \ce{R} groups.
    \begin{equation*}
        \ce{PMe3}
        < \ce{P(Ph3)3}
        < \ce{P($t${-}Bu)3}
    \end{equation*}
    \begin{itemize}
        \item Describes the sterics of the phosphine.
        \item Lists the cone angles of several common phosphines.
    \end{itemize}
    \item \ce{AsR3} (arsenes):
    \begin{itemize}
        \item Isoelectronic/isolobal to phosphines.
        \item Not as common, but typically better $\pi$ acceptors.
    \end{itemize}
    \item N-heterocyclic carbenes:
    \begin{figure}[H]
        \centering
        \schemestart
            \chemfig{R-N=^[:-60]-=^[:60]N-R}
            \arrow{0}[,0]\+{,,6.7mm}
            \chemfig{H-[:30](=[2]O)-[:-30]H}
            \arrow
            \chemfig{[:18]*5(-=N(-R)-(-H)=N(-R)=)}
            \arrow{->[\small\ce{-H+}]}
            \chemfig{[:18]*5(=-NR-\charge{90=\:}{}-RN-)}
        \schemestop
        \caption{Synthesis of N-heterocyclic carbenes.}
        \label{fig:NHeterocyclicCarbeneSynthesis}
    \end{figure}
    \begin{itemize}
        \item A new class of ligands that are extremely strong donors.
        \item Many resonance structures.
        \item A very strong $\sigma$ donor and a very weak $\pi$ acceptor. Essentially pure $\sigma$ donors.
        \begin{itemize}
            \item $\sigma$ donor properties: Orignate from the lone pair on the carbon.
            \item $\pi$ acceptor properties: These exist in principle since the carbon has an empty $p$ orbital, but in practice, it's $\pi$ acceptor abilities are very weak since the carbon $p$ orbital is donated into by the lone pairs on the neighboring nitrogens. Thus, N-heterocyclic carbenes aren't really $\pi$ acceptors.
            \item $\pi$ donor properties: Nonexistent.
        \end{itemize}
        \item Stronger $\sigma$ donors than phosphines.
    \end{itemize}
    \item Other $\pi$-bound ligands.
    \item \ce{CO2} / \ce{CS2}.
    \begin{figure}[h!]
        \centering
        \begin{subfigure}[b]{0.3\linewidth}
            \centering
            \chemfig{M?-[:30]O-[6]C?=[:-60]O}
            \caption{$\eta^2$ bonding.}
            \label{fig:CO2-M-bondinga}
        \end{subfigure}
        \begin{subfigure}[b]{0.3\linewidth}
            \centering
            \begin{tikzpicture}
                \node{
                    \chemfig{M-C(-[:60]O-[,,,,dashed]M)(-[:-60]O-[,,,,dashed]M)}
                };
                \draw [dashed] (0.18,0.56) arc (145:215:1);
            \end{tikzpicture}
            \caption{$\kappa^1$ bonding.}
            \label{fig:CO2-M-bondingb}
        \end{subfigure}
        \begin{subfigure}[b]{0.3\linewidth}
            \centering
            \chemfig{M-O=C=O}
            \caption{End-on bonding.}
            \label{fig:CO2-M-bondingc}
        \end{subfigure}
        \caption{Bonding of \ce{CO2} to metal centers.}
        \label{fig:CO2-M-bonding}
    \end{figure}
    \begin{itemize}
        \item There's a lot of interest in bonding \ce{CO2} to transition metals in terms of catalytic reduction/functionalization of \ce{CO2}.
        \item What is $\kappa^1$ bonding?
    \end{itemize}
    \item Aldehydes / ketones.
    \begin{itemize}
        \item Binding $\eta^2$ and end-on through the oxygen.
    \end{itemize}
\end{itemize}



\section{Lecture 11: Hydrides and Silyls}
\begin{itemize}
    \item \marginnote{4/21:}\ce{M-SiR3} (silyl).
    \begin{itemize}
        \item $1\,\e[-]$ X-type donor.
        \item Closely related to alkyls, but with some important differences (esp. with regard to synthesis and properties).
        \item Important role in a number of catalytic processes (esp. hydrosilylations).
        \item First synthesis reported by Wilkinson in 1956.
        \begin{itemize}
            \item He reacted \ce{Fp- ->[TMSCl] CpFe(CO)2(SiMe3)}.
        \end{itemize}
        \item The bonding is more covalent than with carbon since silicon is more electropositive.
    \end{itemize}
    \item Synthesis of silyls:
    \begin{enumerate}
        \item Electrophilic:
        \begin{itemize}
            \item Mid to late TM's.
            \item Note that $\beta$-\ce{H} elimination is not a problem typically as \ce{Si=CR2} bonds are not stable (unless you add a lot of steric bulk).
            \begin{itemize}
                \item Thus, \ce{SiMe3} does not undergo $\beta$-\ce{H} elimination.
            \end{itemize}
        \end{itemize}
        \item Oxidative addition: \ce{L_nM^0 + HSiR3 <=> L_nMH(SiR3)}.
        \begin{itemize}
            \item Almost always occurs through \ce{Si-H} bonds.
            \item \ce{C-H} addition can be quite hard, but \ce{Si-H} addition can be quite easy.
            \begin{itemize}
                \item Thus, for example, \ce{HSiR2Cl} will add through the \ce{Si-H} bond, not the \ce{Si-Cl} bond since the latter are thermodynamically strong.
            \end{itemize}
            \item \ce{Si-H} bonds form and break in a mobile and dynamic equilibrium.
            \begin{itemize}
                \item The larger orbital radius of silicon helps form the $\sigma$ adducts.
            \end{itemize}
        \end{itemize}
        \item Nucleophilic: \ce{SiR3-} is very rare, so this is uncommon.
        \begin{itemize}
            \item There are some examples, though.
            \begin{itemize}
                \item You can make \ce{LiSiMePh2} in-situ.
                \item You can also make \ce{(THF)3LiSi(TMS)3} and then crystalize it for storage (this compound is known as super silyl).
                \item Lastly, \ce{Cp2ZrCl2 + LiSi(TMS)3 -> Cp2ZrCl(Si(TMS)3)}.
            \end{itemize}
        \end{itemize}
    \end{enumerate}
    \item \ce{M-H} (hydrides and \ce{H2} adducts).
    \begin{itemize}
        \item Pure (and strong) $\sigma$ donors and strong field ligands.
        \item Important intermediates in
        \begin{enumerate}[label={\alph*)}]
            \item Olefin/alkyne hydrogenations.
            \item Isomerizations/polymerizations of olefins.
            \item Hydroformylations.
            \item Decarbonylations.
            \item Hydrogen/deuterium (H/D) exchange.
        \end{enumerate}
        \item Many important structures:
        \begin{itemize}
            \item For example, there are hydrides supported by phosphine ligands, \emph{trans} to carbonyls, \emph{cis} to carbonyls, Cp metal hydrides, and binary hydrides such as \ce{ReH9^2-} (which is face capped trigonal prismatic).
            \item Bridging hydrides: \ce{M-H-M}, \ce{MHM2}, and interstitial hydrides (which is a hydrogen surrounded octahedrally by metals in a solid). Interstitial hydrides are important in solid state hydrogenation catalysts (such as \ce{Pd} / \ce{H2}).
        \end{itemize}
    \end{itemize}
    \item Bond properties of hydrides:
    \begin{itemize}
        \item Hydrides are small compared to typical ligands.
        \begin{itemize}
            \item Thus, for example, \ce{ML4H} will not be trigonal bipyramidal but almost tetrahedral, with the equatorial ligands bending downwards toward the hydride.
            \item The distance between the equatorial ligand plane in \ce{ML5} and \ce{ML4H} is denoted by $d$.
        \end{itemize}
        \item Jim Ibers (emeritus professor at Northwestern) studied various compounds and their $d$ values vs. the \emph{trans} ligand.
        \begin{itemize}
            \item Several examples are listed.
            \item What he found is that bigger ligands make more tetrahedral structures.
        \end{itemize}
        \item The most tetrahedral structures show up as $T_d$ without the hydride in X-ray diffraction since X-rays scatter off of electrons and hydrides have basically none.
        \begin{itemize}
            \item Thus, you need neutron diffraction (which requires really big solid state crystals).
        \end{itemize}
    \end{itemize}
    \item Electronegativity trends in hydrides:
    \begin{itemize}
        \item In general: \chemfig{\charge{90[yshift=1mm]=$\oplus$}{M}-\charge{90[yshift=1mm]=$\ominus$}{H}}
        \item Early metal hydrides are very polar and anionic; late transition metal hydrides are more covalent.
        \item Hydrides can react like \ce{H+}, \ce{H*}, and \ce{H-}.
        \item Electronegativity trends can influence reactivity.
        \item Hydries acting as acids:
        \begin{itemize}
            \item \ce{HCo(CO)4 <=> H+ + Co(CO)4-}, $\pKa<0$ (favorable).
            \item \ce{HCo(CN)5^3- <=> H+ + Co(CN)5^4-}, $\pKa>20$ (unfavorable).
            \item Think about the electronegativity and how electron rich or poor a metal center to determine whether or not it will want to ionize.
        \end{itemize}
        \item Hydrides acting as classic hydrides:
        \begin{itemize}
            \item \ce{Cp^*2ZrH2 ->[HCl] H2 + Cp^*2ZrHCl}.
            \item \ce{Cp^*2ZrH2 ->[2 O=CH2] Cp^*2Zr(OCH3)2}.
        \end{itemize}
        \item Hydrides acting somewhere in-between these two extremes (actin acidic and hydritic):
        \begin{itemize}
            \item \ce{Cp2MoH2 ->[LDA] [Cp2MoH^- Li]4}.
            \item Note that LDA is a strong base.
            \item \ce{Cp2MoH2 <=>[H+] Cp2MoH3+}.
            \item Note that \ce{H+} is just a generic acid, and the product is an \ce{H2} adduct.
        \end{itemize}
        \item Radical:
        \begin{itemize}
            \item \ce{M-H -> M* + H*}.
            \item The classic example of this is \ce{(CO)5Mn-H} which can homolyze pretty easily.
        \end{itemize}
    \end{itemize}
    \item Synthesis of hydrides (the first four are fundamental reactions, but the latter ones are also important):
    \begin{enumerate}
        \item \ce{H2} directly:
        \begin{itemize}
            \item \ce{CoS + [CO + H2] ->[$\SI{200}{atm}$][Cu / $\SI{200}{\celsius}$] HCo(CO)4 + CuS}.
            \begin{itemize}
                \item \ce{[CO + H2]} is a mixture of the two gasses known as Syn gas.
                \item The key to making this reaction proceed is just temperature and pressure. Catalysts can also help.
            \end{itemize}
            \item \ce{FeI2 + PF3 + H2 ->[$\SI{100}{atm}$][Zn / $\SI{250}{\celsius}$] H2Fe(PF3)4 + ZnI2}.
            \begin{itemize}
                \item Zinc serves as a reductant here.
            \end{itemize}
            \item Other examples exist, but no more are shown.
        \end{itemize}
        \item Protonation of metal anions:
        \begin{itemize}
            \item \ce{Ru(CO)5 ->[Na][-CO]Na2Ru(CO)4 ->[HA] H2Ru(CO)4}.
            \item Similar chemistry exists with \ce{Os} and \ce{Fe}.
        \end{itemize}
        \item Oxidative addition across a \ce{M-M} bond.
        \begin{itemize}
            \item \ce{Cp(CO)3Cr-Cr(CO)3Cp ->[H2] 2CpCr(CO)3H}.
            \item \ce{Os3(CO)12 ->[H2][$\SI{80}{atm}$] H2Os(CO)4}.
        \end{itemize}
        \item Oxidative addition to a single metal center.
        \begin{itemize}
            \item \ce{IrCl(CO)(PPh3)2 ->[H2] IrClH2(CO)(PPh3)2}.
            \item \ce{HCo(N2)(PPh3)3 ->[-N2][H2] H3Co^{III}(PPh3)3}
        \end{itemize}
        \item Hydrogenations of \ce{M-L} bonds with \ce{H2}.
        \begin{itemize}
            \item \ce{M-R + H2 -> [MRH2]^{$\ddagger$} -> M-H + R-H}.
            \item \ce{M-OR + H2 -> M-H + HOR}.
            \begin{itemize}
                \item Much harder to do since the \ce{M-O} bond is typically more ionic, and thus stronger.
            \end{itemize}
            \item Example: \ce{Cp^*2Ta(CH3)4 ->[H2] Cp^*2TaH4 + 4CH4}.
            \begin{itemize}
                \item This reaction must proceed through $\sigma$-bond metathesis.
            \end{itemize}
        \end{itemize}
        \item Electrophilic metal centers with hydride reagents.
        \begin{itemize}
            \item \ce{L3IrCl2 ->[LAlH][THF] IrH3L3 + H3Al-THF}.
            \item \ce{Cp2ZrCl2 ->[BH4-] Cp2Zr(BH4)2 ->[NR3] [Cp2ZrH2]2 + R3N-BH3}.
        \end{itemize}
        \item $\beta$-\ce{H} elimination.
        \begin{figure}[h!]
            \centering
            \begin{subfigure}[b]{0.9\linewidth}
                \centering
                \schemestart
                    \chemfig{M-X-CHR_2}
                    \arrow
                    \chemfig{M(-[1]H)-[7]\phantom{i}-[1,0.4,,,white]X=[5]R_2C}
                \schemestop
                \caption{General case.}
                \label{fig:beta-H-synthesizeMetalHydridesa}
            \end{subfigure}
            \vspace{1em}
            \begin{subfigure}[b]{0.49\linewidth}
                \centering
                \schemestart
                    \chemfig{M-[:30]-[:-30]}
                    \arrow
                    \chemfig{M(-[1]H)-[7]\phantom{i}-[1,0.4,,,white]=[5]}
                \schemestop
                \caption{Simplest case.}
                \label{fig:beta-H-synthesizeMetalHydridesb}
            \end{subfigure}
            \begin{subfigure}[b]{0.49\linewidth}
                \centering
                \schemestart
                    \chemfig{M-[:30]O-[:-30]CH_3}
                    \arrow
                    \chemfig{M(-[1]H)-[7]\phantom{i}-[1,0.4,,,white]O=[5](-[4]H)(-[6]H)}
                \schemestop
                \caption{Alternate case.}
                \label{fig:beta-H-synthesizeMetalHydridesc}
            \end{subfigure}
            \caption{$\beta$-\ce{H} elimination synthesis of metal hydrides.}
            \label{fig:beta-H-synthesizeMetalHydrides}
        \end{figure}
        \begin{itemize}
            \item The relative direction of the equilibrium depends on the system, but this can definitely happen.
            \item Goes over how to isolate osmium from osmium ore.
        \end{itemize}
    \end{enumerate}
    \item Non-classical and paramagnetic hydrides.
    \begin{figure}[h!]
        \centering
        \begin{subfigure}[b]{0.3\linewidth}
            \centering
            \chemfig{\ce{L_n}M(-[1]H)(-[7]H)}
            \caption{General dihydride.}
            \label{fig:nonclassicalHydridea}
        \end{subfigure}
        \begin{subfigure}[b]{0.3\linewidth}
            \centering
            \chemfig{\ce{L_n}M-[,,,,{stealth-}]\phantom{i}-[2,0.3,,,white]H-[6]H}
            \caption{General \ce{H2} adduct.}
            \label{fig:nonclassicalHydrideb}
        \end{subfigure}
        \caption{Non-classical hydrides.}
        \label{fig:nonclassicalHydride}
    \end{figure}
    \begin{itemize}
        \item \ce{L_nMH2} could be a dihydride or an \ce{H2} adduct (or some resonance structure in between). The latter is a non-classical hydride.
        \item We distinguish between the two with NMR spectroscopy since \ce{{}^1H} hydrides are usually upfield ($-1$ to $\SI{-15}{ppm}$).
        \item A \ce{T1} relaxation also provides the \ce{H-H} distance.
        \item Lastly, \ce{H1D} isotopologues give you a \ce{H-D} coupling constant, which can be correlated to bond distances.
        \item If you have paramagnetic complexes, this is much harder, so you have to use special techniques such as ENDOR  spectroscopy.
    \end{itemize}
\end{itemize}




\end{document}